---
title: "Week 10 Lab: Interactions"
subtitle: "Applications from the lecture — work through the code and check that it runs"
author: "Firstname Lastname"
date: last-modified
date-format: MMMM D, YYYY
format: html
toc: true
toc-depth: 3
code-fold: true
code-tools: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# Setup {#sec-setup}

Before running any analysis, we load the packages we need and define a few
shared colour constants that keep plots visually consistent.
`brms` fits Bayesian regression models via Stan; `bayesplot` produces
MCMC diagnostics; `patchwork` assembles multi-panel figures.

```{r}
#| label: setup
#| code-fold: false
start_time <- Sys.time() # used to measure execution time at the end

library(tidyverse) # data manipulation and ggplot2
library(brms) # Bayesian regression via Stan
library(bayesplot) # MCMC diagnostic plots
library(patchwork) # composing multi-panel figures

# Shared colour palette — reused in every plot so figures look consistent
col_1 <- "steelblue"
col_2 <- "#E07B39"
col_3 <- "#7B3F9E"
col_4 <- "goldenrod3"
chain_cols <- c(col_1, col_2, col_3, col_4) # one colour per MCMC chain
```

---

# Part II: Application — Interaction `brms` Model {#sec-part1}

## The Research Question {#sec-part1-intro}

Last week we established that women's parliamentary representation is associated
with higher health spending, after controlling for GDP per capita and regime
type as an additive intercept shift.
But the additive model assumes the **slope** of the women-in-parliament
association is the same for both presidential and parliamentary systems.

Part I relaxes this assumption:

> **Does the association between female legislator share and health spending vary by regime type?**

**Legislative composition theory**: parliamentary systems concentrate budget
authority in the legislature — strong party discipline, parliamentary
sovereignty, budget processes rooted in the legislature.
Presidential systems divide power across branches, potentially weakening the
translation of legislative preferences into policy.
This predicts a **steeper slope** for women in parliament in parliamentary systems.

## DGP: Synthetic Data {#sec-part1-dgp}

We use the **same synthetic dataset** as Week 9, reproduced here so this lab
runs standalone.
The data-generating process already encodes the interaction: the true women-in-parliament
slope is 0.08 in presidential systems and 0.18 in parliamentary systems.

```{r}
#| label: dgp
set.seed(2026) # same seed as Week 9 — identical dataset
n <- 120

# ── Population-level parameters (the ground truth) ──────────────────────────
beta_0 <- 2 # intercept (presidential baseline)
beta_women_pres <- 0.08 # women_parl slope in presidential systems
beta_gdp <- 0.20 # GDP per capita partial association
beta_reg <- 1.5 # parliamentary intercept shift
beta_interact <- 0.10 # extra women_parl slope in parliamentary systems
sigma <- 1.5 # residual standard deviation

# ── Data-generating process ──────────────────────────────────────────────────
gdp_pc <- rnorm(n, mean = 18, sd = 7)
women_parl <- 15 + 0.9 * gdp_pc + rnorm(n, 0, 5)
reg_type <- rbinom(n, 1, prob = 0.55) # 1 = parliamentary, 0 = presidential

health_spend <- beta_0 +
  beta_gdp * gdp_pc +
  beta_women_pres * women_parl +
  beta_reg * reg_type +
  beta_interact * women_parl * reg_type + # steeper slope in parliamentary
  rnorm(n, 0, sigma)

df <- tibble(gdp_pc, women_parl, reg_type, health_spend) |>
  mutate(system = if_else(reg_type == 1, "Parliamentary", "Presidential"))
```

True women-in-parliament slopes:

| System | Slope |
|---|---|
| Presidential (`reg_type = 0`) | **0.08** |
| Parliamentary (`reg_type = 1`) | **0.08 + 0.10 = 0.18** |

The interaction coefficient **$\beta_3 = 0.10$** captures *how much steeper*
the parliamentary slope is relative to the presidential slope.

## Step 1: DAG and Estimand {#sec-part1-step1}

The DAG from Week 9 shows regime type having a **direct arrow**
into health spending (intercept shift, captured last week) and a **moderation
role** on the women-in-parliament → health-spending edge (the interaction we
estimate today).

Because we already control for GDP per capita (the confounder), no additional
adjustment is needed.
The interaction term itself is the new quantity of interest.

**Estimand 1**: $\partial\mu/\partial X \mid Z=0$ — women-in-parliament slope
in *presidential* systems.

**Estimand 2**: $\partial\mu/\partial X \mid Z=1$ — women-in-parliament slope
in *parliamentary* systems.

**Estimand 3**: $\beta_3$ — the slope difference; how much more does women's
representation associate with health spending in parliamentary vs. presidential
systems?

## Step 2: Data Summary {#sec-part1-step2}

We start with the marginal distribution of health spending by system type,
then examine the joint scatter with separate OLS lines per group and a
tercile-based boxplot comparison.

```{r}
#| label: fig-outcome-dist-interact
#| echo: true
#| fig-height: 3.0
#| dpi: 500
#| fig-cap: "Marginal distribution of health spending by regime type. Parliamentary systems show a higher average — the intercept shift β₂ captured in the additive model from Week 9."
df |>
  ggplot(aes(x = health_spend, fill = system, colour = system)) +
  geom_density(alpha = 0.45, linewidth = 0.8) +
  scale_fill_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2)
  ) +
  scale_colour_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2)
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Health Spending (% GDP)",
    y = "Density",
    title = "Distribution of Health Spending by Regime Type",
    subtitle = "Parliamentary systems spend more on average",
    fill = NULL,
    colour = NULL
  )
```

::: {.callout-note}
## Alternative implementation

```r
df |>
  ggplot(aes(x = system, y = health_spend, fill = system)) +
  geom_violin(alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.35, size = 1.2) +
  scale_fill_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2), guide = "none"
  ) +
  theme_bw() +
  labs(title = "Health spending by regime type — violin view",
       x = NULL, y = "Health Spending (% GDP)")
```

:::

The scatter below puts women in parliament on the x-axis and draws separate OLS
lines for each system type.
If the lines have visibly different slopes, that is preliminary evidence for
an interaction.
The boxplot facets women in parliament into terciles and shows whether the gap
between systems widens with more female legislators — the visual signature of
$\beta_3 > 0$.

```{r}
#| label: fig-interact-summary
#| echo: true
#| code-fold: true
#| fig-height: 3.6
#| dpi: 500
#| fig-cap: "Left: scatter with separate OLS lines per system — slopes appear different. Right: health spending by women-in-parliament tercile and system type — if the gap widens with more women in parliament, that is the visual signature of a positive interaction."
p_scatter_int <- df |>
  ggplot(aes(x = women_parl, y = health_spend, colour = system)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(
    aes(group = system),
    method = "lm",
    se = FALSE,
    linewidth = 1.2
  ) +
  scale_colour_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2),
    name = "System"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Spending (% GDP)",
    title = "Separate Slopes by Regime Type",
    subtitle = "OLS line per system"
  )

df_interact_plot <- df |>
  mutate(
    wp_tercile = cut(
      women_parl,
      breaks = quantile(women_parl, c(0, 1 / 3, 2 / 3, 1)),
      labels = c("Low\n(<25%)", "Mid\n(~33%)", "High\n(>40%)"),
      include.lowest = TRUE
    )
  )

p_box_int <- df_interact_plot |>
  ggplot(aes(x = wp_tercile, y = health_spend, fill = system)) +
  geom_boxplot(
    alpha = 0.7,
    position = position_dodge(width = 0.75),
    width = 0.6
  ) +
  scale_fill_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2),
    name = "System"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Women in Parliament Tercile",
    y = "Health Spending (% GDP)",
    title = "Gap Across Systems by Women-in-Parliament Level",
    subtitle = "Widening gap = signature of a positive interaction"
  )

p_scatter_int | p_box_int
```

::: {.callout-note}
## Alternative implementation

```r
df |>
  mutate(tercile = cut(women_parl,
    breaks = quantile(women_parl, c(0, 1/3, 2/3, 1)),
    labels = c("Low WP", "Mid WP", "High WP"),
    include.lowest = TRUE)) |>
  ggplot(aes(x = tercile, y = health_spend, fill = system)) +
  geom_boxplot(position = position_dodge(0.75), alpha = 0.7) +
  scale_fill_manual(values = c("Parliamentary" = col_1, "Presidential" = col_2)) +
  theme_bw() +
  labs(title = "Spending gap across systems — tercile view",
       x = "Women in Parliament Tercile", y = "Health Spending (% GDP)", fill = NULL)
```

:::

## Step 3: Formal Model Specification {#sec-part1-step3}

The interaction model augments the additive model with a product term:

$$
Y_i \sim \text{Normal}(\mu_i,\, \sigma)
$$

$$
\mu_i = \alpha + \beta_1 X_i + \beta_2 D_i + \beta_3 (X_i \cdot D_i) + \beta_Z Z_i
$$

where $X$ = women in parliament (%), $D$ = regime type (1 = parliamentary),
$Z$ = GDP per capita.

The two conditional marginal effects are now *functions* of the moderator:

$$
\frac{\partial\mu}{\partial X} = \beta_1 + \beta_3 D \qquad
\frac{\partial\mu}{\partial D} = \beta_2 + \beta_3 X
$$

**Priors**:

$$
\alpha \sim \text{Normal}(10, 5) \qquad
\beta_1, \beta_2, \beta_Z \sim \text{Normal}(0, 2) \qquad
\beta_3 \sim \text{Normal}(0, 2) \qquad
\sigma \sim \text{Exponential}(1)
$$

The prior for $\beta_3$ is centred at zero — our prior expectation is *no moderation*,
given the data and our prior.
This is appropriately humble: we should not encode strong prior beliefs about
the direction of an interaction before seeing the evidence.

We fit two competing models:

- **`fit_multi`** (no interaction): assumes the women-in-parliament slope is the
  same for all system types — a testable restriction.
- **`fit_interact`** (interaction): relaxes this restriction.
  If `fit_interact` fits substantially better via LOO-IC, we have evidence that the
  restriction is false.

## Step 4: Prior Predictive Check {#sec-part1-step4}

For the interaction model, we draw regression lines from the joint prior —
one bundle per regime type.
The check asks: *does the prior allow a wide enough range of slope differences,
including no moderation (parallel lines), strong moderation (strongly diverging
lines), and implausible configurations?*

```{r}
#| label: fig-prior-ppc-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.4
#| dpi: 500
#| fig-cap: "Prior predictive spaghetti: 150 regression lines drawn from the joint prior, separately for parliamentary (blue) and presidential (orange) systems. The prior is wide but not implausible — it allows no interaction, moderate interaction, and strong interaction."
set.seed(2026)
n_pp2 <- 150
wp_seq <- seq(10, 55, length.out = 60)
gdp_mean <- mean(df$gdp_pc)

prior_lines_int <- tibble(
  sim = 1:n_pp2,
  b0 = rnorm(n_pp2, 10, 5), # intercept draws — Normal(10, 5)
  b_wp = rnorm(n_pp2, 0, 2), # women_parl slope draws — Normal(0, 2)
  b_reg = rnorm(n_pp2, 0, 2), # regime-type main effect draws — Normal(0, 2)
  b_int = rnorm(n_pp2, 0, 2), # interaction term draws — Normal(0, 2)
  b_gdp = rnorm(n_pp2, 0, 2) # GDP slope draws — Normal(0, 2)
) |>
  expand_grid(women_parl = wp_seq, reg_type = c(0L, 1L)) |>
  mutate(
    y_hat = b0 +
      b_wp * women_parl +
      b_reg * reg_type +
      b_int * women_parl * reg_type + # slope shifts by b_int in parliamentary
      b_gdp * gdp_mean,
    system = if_else(reg_type == 1, "Parliamentary", "Presidential")
  )

prior_lines_int |>
  ggplot(aes(
    x = women_parl,
    y = y_hat,
    group = interaction(sim, system),
    colour = system
  )) +
  geom_line(alpha = 0.12, linewidth = 0.5) +
  scale_colour_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2),
    name = "System"
  ) +
  coord_cartesian(ylim = c(-5, 30)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Women in Parliament (%)",
    y = "Predicted Health Spending (% GDP)",
    title = "Prior Predictive Spaghetti: Interaction Model",
    subtitle = "150 lines per system — prior is non-dogmatic about moderation"
  )
```

## Step 5: Fit Both Models {#sec-part1-step5}

We use the `*` operator in the `brm()` formula.
`women_parl * reg_type` is shorthand for `women_parl + reg_type + women_parl:reg_type`.
Always use `*`, not just the colon form (`:`).
The colon-only form omits the main effects, which forces $\beta_1 = 0$ and
$\beta_2 = 0$ — usually a false and arbitrary constraint.

We fit `fit_multi` (additive, no interaction) as the comparison baseline, then
`fit_interact` (the model of interest).

```{r}
#| label: fit-multi
#| output: false
# Additive multivariate model — comparison baseline from Week 9
fit_multi <- brm(
  health_spend ~ women_parl + gdp_pc + reg_type,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2), class = "b", coef = "women_parl"),
    prior(normal(0, 2), class = "b", coef = "gdp_pc"),
    prior(normal(0, 2), class = "b", coef = "reg_type"),
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

```{r}
#| label: fit-interact
#| echo: true
#| output: false
# Interaction model — allows the women_parl slope to differ by regime type
# * expands to: women_parl + reg_type + women_parl:reg_type
fit_interact <- brm(
  health_spend ~ women_parl * reg_type + gdp_pc,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2), class = "b", coef = "women_parl"), # presidential slope
    prior(normal(0, 2), class = "b", coef = "reg_type"), # parliamentary shift
    prior(normal(0, 2), class = "b", coef = "gdp_pc"), # GDP partial effect
    prior(normal(0, 2), class = "b", coef = "women_parl:reg_type"), # slope difference
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

::: {.callout-note}
## Alternative implementation

```r
# Confirm priors actually applied to the interaction model
prior_summary(fit_interact)
```

:::

## Step 6: Diagnostics and Posterior Predictive Check {#sec-part1-step6}

We run the same convergence checks as before.
The traceplots and PPC are shown side-by-side for the interaction model.

```{r}
#| label: fig-diag-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Left: traceplots for the five parameters of the interaction model — all chains mix well. Right: posterior predictive check — model predictions envelope the observed distribution."
p_trace_int <- mcmc_trace(
  fit_interact,
  pars = c(
    "b_Intercept",
    "b_women_parl",
    "b_reg_type",
    "b_women_parl:reg_type",
    "b_gdp_pc"
  ),
  facet_args = list(ncol = 2)
) +
  scale_colour_manual(values = chain_cols) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90"),
    legend.position = "none"
  ) +
  labs(title = "Traceplots: Interaction Model")

p_ppc_int <- pp_check(fit_interact, ndraws = 50) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "Posterior Predictive Check")

p_trace_int | p_ppc_int
```

::: {.callout-note}
## Alternative implementation

```r
mcmc_rank_hist(
  fit_interact,
  pars = c("b_Intercept", "b_women_parl",
           "b_reg_type", "b_women_parl:reg_type", "b_gdp_pc")
) +
  labs(title = "Rank histograms: interaction model convergence check")
```

:::

## Step 7: Interpreting Results and Posterior Marginal Effects {#sec-part1-step7}

Inspect the summary output.
The key quantity is `b_women_parl:reg_type` — the interaction coefficient $\beta_3$.
Its posterior mean should be near 0.10 (the true DGP value), given the data and prior.

```{r}
#| label: interact-summary
#| echo: true
summary(fit_interact)
```

::: {.callout-note}
## Alternative implementation

```r
as_draws_df(fit_interact) |>
  mutate(
    slope_presidential  = b_women_parl,
    slope_parliamentary = b_women_parl + `b_women_parl:reg_type`
  ) |>
  summarise(across(
    c(slope_presidential, slope_parliamentary),
    list(mean  = mean,
         q2.5  = \(x) quantile(x, 0.025),
         q97.5 = \(x) quantile(x, 0.975))
  ))
```

:::

### Posterior Line Clouds

For each MCMC draw we compute the intercept and slope for each system type
by simple arithmetic and draw the resulting line.
The bundle of parliamentary lines (blue) should be steeper than the presidential
bundle (orange).

```{r}
#| label: fig-posterior-lines-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Posterior regression line clouds: 200 MCMC draws per system type. The blue bundle (parliamentary) is visibly steeper — posterior evidence for β₃ > 0."
set.seed(2026)
post_draws_int <- as_draws_df(fit_interact) |> slice_sample(n = 200)
gdp_mean <- mean(df$gdp_pc)

df |>
  ggplot(aes(x = women_parl, y = health_spend, colour = system)) +
  # Presidential lines: intercept = b_Intercept + b_gdp_pc * gdp_mean,
  #                     slope     = b_women_parl
  geom_abline(
    data = post_draws_int,
    aes(
      intercept = b_Intercept + b_gdp_pc * gdp_mean,
      slope = b_women_parl
    ),
    colour = col_2,
    alpha = 0.05,
    linewidth = 0.4
  ) +
  # Parliamentary lines: intercept = b_Intercept + b_reg_type + b_gdp_pc * gdp_mean,
  #                      slope     = b_women_parl + b_women_parl:reg_type
  geom_abline(
    data = post_draws_int,
    aes(
      intercept = b_Intercept + b_reg_type + b_gdp_pc * gdp_mean,
      slope = b_women_parl + `b_women_parl:reg_type`
    ),
    colour = col_1,
    alpha = 0.05,
    linewidth = 0.4
  ) +
  geom_point(alpha = 0.5, size = 1.8) +
  scale_colour_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2),
    name = "System"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Spending (% GDP)",
    title = "Posterior Line Clouds by Regime Type",
    subtitle = "200 MCMC draws per group — blue bundle (parliamentary) is steeper"
  )
```

::: {.callout-note}
## Alternative implementation

```r
# brms built-in: automatic interaction plot
plot(conditional_effects(
  fit_interact,
  effects = "women_parl:reg_type"
))
```

:::

### Posterior Conditional Marginal Effects

The conditional marginal effect of women in parliament for a given system type
is a function of $\beta_1$ and $\beta_3$:

$$
\text{ME}_{\text{presidential}}  = \beta_1 \qquad
\text{ME}_{\text{parliamentary}} = \beta_1 + \beta_3
$$

Because we have full posterior draws of both $\beta_1$ and $\beta_3$, we compute
these by arithmetic — no approximation or delta method needed.
The resulting distributions are posterior statements, given the data and prior.

```{r}
#| label: fig-marginal-effects-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.4
#| dpi: 500
#| fig-cap: "Posterior distributions of the conditional marginal effect of women in parliament on health spending by system type. The parliamentary slope (~0.18) and presidential slope (~0.08) should be well separated — posterior evidence for β₃ > 0, given the data and prior."
post_all_int <- as_draws_df(fit_interact)

me_df <- post_all_int |>
  transmute(
    `Presidential` = b_women_parl, # ME at D = 0
    `Parliamentary` = b_women_parl + `b_women_parl:reg_type` # ME at D = 1
  ) |>
  pivot_longer(
    everything(),
    names_to = "system",
    values_to = "marginal_effect"
  ) |>
  mutate(
    system = factor(system, levels = c("Presidential", "Parliamentary"))
  )

# P(β₃ > 0 | data): direct posterior probability — not a p-value
prob_b3_pos <- mean(post_all_int$`b_women_parl:reg_type` > 0)

me_df |>
  ggplot(aes(x = marginal_effect, fill = system, colour = system)) +
  geom_density(alpha = 0.55, linewidth = 0.8) +
  geom_vline(
    xintercept = 0,
    colour = "gray40",
    linetype = "dotted",
    linewidth = 0.9
  ) +
  scale_fill_manual(
    values = c("Presidential" = col_2, "Parliamentary" = col_1),
    name = NULL
  ) +
  scale_colour_manual(
    values = c("Presidential" = col_2, "Parliamentary" = col_1),
    name = NULL
  ) +
  annotate(
    "text",
    x = max(me_df$marginal_effect) * 0.7,
    y = Inf,
    vjust = 1.5,
    size = 3.5,
    colour = col_3,
    label = paste0("P(β_3 > 0 | data) = ", round(prob_b3_pos, 3))
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = expression(partialdiff * mu / partialdiff ~ "Women in Parliament (%)"),
    y = "Posterior Density",
    title = "Posterior Marginal Effects: Women in Parliament -> Health Spending",
    subtitle = "Parliamentary slope ~0.18  |  Presidential slope ~0.08  |  Exact by arithmetic"
  )
```

::: {.callout-note}
## Alternative implementation

```r
# Direct posterior summary of the interaction term
as_draws_df(fit_interact) |>
  summarise(
    mean_b3  = mean(`b_women_parl:reg_type`),
    q2.5     = quantile(`b_women_parl:reg_type`, 0.025),
    q97.5    = quantile(`b_women_parl:reg_type`, 0.975),
    prob_pos = mean(`b_women_parl:reg_type` > 0)
  )
```

:::

**Reading the figure**: the two density curves should be separated.
`P(β₃ > 0 | data)` near 1.0 means we are nearly certain, given the data and prior,
that the interaction is positive — parliamentary systems show a steeper women-in-parliament–health-spending
association.
This is a direct posterior probability statement, not a p-value.

## Step 8: LOO-IC Model Comparison {#sec-part1-step8}

We compare `fit_multi` (no interaction) against `fit_interact` (interaction) using LOO-IC.
Because we generated the data with an interaction, `fit_interact` should dominate `fit_multi`.

```{r}
#| label: loo-interact
#| echo: true
fit_multi <- add_criterion(fit_multi, "loo")
fit_interact <- add_criterion(fit_interact, "loo")
loo_compare(fit_multi, fit_interact)
```

::: {.callout-note}
## Alternative implementation

```r
# WAIC as alternative to LOO-IC — should rank models identically
fit_multi    <- add_criterion(fit_multi,    "waic")
fit_interact <- add_criterion(fit_interact, "waic")
loo_compare(fit_multi, fit_interact, criterion = "waic")
```

:::

**Reading the output**:

- The model listed first has the highest elpd — it is the better predictive model.
- `elpd_diff` = 0 for the winner; all other models show a negative difference
  (i.e., they predict worse).
- When `|elpd_diff| > 2 × se_diff`, the evidence strongly favours the winner.

**Remaining limitations**:

1. The linear interaction assumes the slope *difference* is constant across all
   values of women in parliament — check with a LOESS diagnostic if needed.
2. Country-year observations within the same country are correlated (assumption A5) —
   a multilevel structure (Week 12) would handle this.
3. Omitted confounders (institutional quality, colonial history, cultural factors)
   may bias the women-in-parliament coefficient.

---

# Session Info and Execution Time {#sec-session}

```{r}
#| label: session-info
#| eval: true
sessionInfo()
```

```{r}
#| label: exec-time
#| eval: true
end_time <- Sys.time()
cat(
  "Total execution time:",
  round(difftime(end_time, start_time, units = "mins"), 1),
  "minutes\n"
)
```
