---
title: "Week 9 Lab: Multivariate Regression"
subtitle: "Applications from the lecture — work through the code and check that it runs"
author: "Firstname Lastname"
date: last-modified
date-format: MMMM D, YYYY
format: html
toc: true
toc-depth: 3
code-fold: true
code-tools: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# Setup {#sec-setup}

Before running any analysis, we load the packages we need and define a few
shared colour constants that keep plots visually consistent.
`brms` fits Bayesian regression models via Stan; `bayesplot` produces
MCMC diagnostics; `patchwork` assembles multi-panel figures.

```{r}
#| label: setup
#| code-fold: false
start_time <- Sys.time() # used to measure execution time at the end

library(tidyverse) # data manipulation and ggplot2
library(brms) # Bayesian regression via Stan
library(bayesplot) # MCMC diagnostic plots
library(patchwork) # composing multi-panel figures

# Shared colour palette — reused in every plot so figures look consistent
col_1 <- "steelblue"
col_2 <- "#E07B39"
col_3 <- "#7B3F9E"
col_4 <- "goldenrod3"
chain_cols <- c(col_1, col_2, col_3, col_4) # one colour per MCMC chain
```

---

# Part II: Application I — Multivariate `brms` Model {#sec-part2}

## DAGs in Quarto: MermaidJS {#sec-dag-mermaid}

One quick way to sketch a DAG directly in a Quarto document — no R package required.
Use a fenced `{mermaid}` block with `graph LR` (left-to-right) or `graph TD` (top-down).

```{mermaid}
graph LR
  GDP["GDP per Capita"] --> WP["Women in Parliament (%)"]
  GDP --> HS["Health Spending (% GDP)"]
  WP --> HS
  RT["Regime Type"] --> HS
```

Nodes take labels in `["quoted text"]`; directed edges are `-->`.
This is enough for a homework submission or a quick theoretical sketch.

## DGP: Synthetic Data (Dataset 1) {#sec-part2-dgp}

This lab uses a **single synthetic dataset** throughout.
Working with synthetic data is pedagogically powerful: we know the true
parameter values, so we can verify whether our models recover them.
This is something you can almost never do with real data — in practice the DGP
is unobserved; we only see the outcomes, never the process that produced them.

The research question for this week:

- Is the share of women in parliament associated with higher public health expenditure, adjusting for GDP per capita and regime type?

The key complication is **confounding**: wealthier countries
simultaneously have more women in parliament *and* spend more on health.
Failing to account for GDP per capita will over-estimate the legislative
representation association.
Week 10 adds a second layer: whether the slope itself differs by regime type
(association heterogeneity), addressed via an interaction term.

```{r}
#| label: dgp
set.seed(2026) # reproducibility: same data every time
n <- 120 # 120 synthetic country-years

# ── Population-level parameters (the ground truth) ──────────────────────────
beta_0 <- 2 # intercept (presidential baseline)
beta_women_pres <- 0.08 # women_parl slope in presidential systems
beta_gdp <- 0.20 # GDP per capita partial association
beta_reg <- 1.5 # parliamentary intercept shift
beta_interact <- 0.10 # extra women_parl slope in parliamentary systems
sigma <- 1.5 # residual standard deviation

# ── Data-generating process ──────────────────────────────────────────────────
gdp_pc <- rnorm(n, mean = 18, sd = 7)
women_parl <- 15 + 0.9 * gdp_pc + rnorm(n, 0, 5)
reg_type <- rbinom(n, 1, prob = 0.55) # 1 = parliamentary, 0 = presidential

health_spend <- beta_0 +
  beta_gdp * gdp_pc +
  beta_women_pres * women_parl +
  beta_reg * reg_type +
  beta_interact * women_parl * reg_type + # steeper slope in parliamentary
  rnorm(n, 0, sigma)

df <- tibble(gdp_pc, women_parl, reg_type, health_spend) |>
  mutate(system = if_else(reg_type == 1, "Parliamentary", "Presidential"))
```

::: {.callout-note}
## Alternative implementation

```r
# OLS check: verify OVB direction without brms
m_biv  <- lm(health_spend ~ women_parl, data = df)
m_full <- lm(health_spend ~ women_parl + gdp_pc + reg_type, data = df)
round(coef(m_biv)["women_parl"] - coef(m_full)["women_parl"], 3)
```

:::

True women-in-parliament slopes:

| System | Slope |
|---|---|
| Presidential (`reg_type = 0`) | **0.08** |
| Parliamentary (`reg_type = 1`) | **0.08 + 0.10 = 0.18** |

Because GDP drives both variables, a bivariate model (no controls)
will produce an estimate well above either true value — that excess is omitted
variable bias (OVB).
The additive multivariate model (Part II here) estimates an *average* slope
across both system types; the interaction model (Week 10 lab) estimates
separate slopes.

## Step 1: Estimand {#sec-part2-step1}

Before writing any code, we state precisely what we are trying to estimate.
Drawing a DAG before analysis keeps this honest: the estimand follows from
the causal graph, not from what the model happens to produce.

Our DAG (presented in the lecture) shows GDP per capita as a **confounder** —
it has causal arrows into both women's parliamentary representation and health
spending.
A confounder opens a "back-door path" that conflates the association we care
about with the spurious GDP-driven correlation.
Conditioning on GDP per capita — including it as a predictor — closes that path.
Regime type (`reg_type`) enters as a **dummy variable** for the intercept shift
(parliamentary systems may have a different baseline level of spending).
Whether it also moderates the women-in-parliament slope is the question we take up in Week 10.

**Estimand (Part II)**: the partial association between women's parliamentary
representation and public health expenditure, adjusting for GDP per capita and
regime type — the average slope across both system types.

*Adjusting for GDP per capita and regime type shifts the estimate toward the true DGP average. Whether the adjustment set is sufficient is a substantive question the data alone cannot answer.*

## Step 2: Data Summary {#sec-part2-step2}

The scatter below shows the raw bivariate association between women's
parliamentary share and health spending.
There is a positive trend, but it conflates the legislative association with
the confounding from GDP.
We will visualise the OVB explicitly in Step 7, after fitting both models.

```{r}
#| label: fig-controls-raw
#| echo: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Raw bivariate scatter of women's parliamentary share against health expenditure. The positive trend conflates the legislative association with the GDP confounder."
df |>
  ggplot(aes(x = women_parl, y = health_spend)) +
  geom_point(colour = col_1, alpha = 0.55, size = 2) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Expenditure (% GDP)",
    title = "Raw Bivariate Association",
    subtitle = "Positive trend — but GDP confounds the relationship"
  )
```

::: {.callout-note}
## Alternative implementation

```r
df |>
  ggplot(aes(x = women_parl, y = health_spend, colour = gdp_pc)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_colour_gradient(low = col_1, high = col_2, name = "GDP pc") +
  theme_bw() +
  labs(title = "Confounding visible: GDP colours the point cloud",
       x = "Women in Parliament (%)", y = "Health Expenditure (% GDP)")
```

:::

## Step 3: Formal Model Specification {#sec-part2-step3}

We write out the full statistical model — likelihood and priors — before
touching `brms`.
This makes our assumptions explicit and contestable.

**Likelihood** (unchanged from Week 8 — still Gaussian):

$$
Y_i \sim \text{Normal}(\mu_i,\, \sigma)
$$

$$
\mu_i = \alpha + \beta_X X_i + \beta_Z Z_i + \beta_D D_i
$$

where $X$ = women in parliament (%), $Z$ = GDP per capita (thousands USD, PPP),
$D$ = regime type (1 = parliamentary, 0 = presidential).
The dummy $D$ gives parliamentary systems a different intercept baseline.

**Priors** (one per unknown parameter — all five must be justified):

$$
\alpha \sim \text{Normal}(10, 5) \qquad \beta_X \sim \text{Normal}(0, 2)
$$

$$
\beta_Z \sim \text{Normal}(0, 2) \qquad \beta_D \sim \text{Normal}(0, 2) \qquad \sigma \sim \text{Exponential}(1)
$$

The prior on $\beta_X$ says: *before seeing data, a 1 pp increase in women's
representation is associated with a change of somewhere between −6 and +6 pp
in health spending (the Normal(0, 2) range at ±3σ is [−6, +6]).*
The prior on $\beta_D$ says: *parliamentary systems may spend anywhere from about 6 pp
less to 6 pp more than presidential systems, holding other predictors constant —
a weakly informative, non-dogmatic prior with no assumption about which system spends more.*
Both are defensible weakly informative priors: broad enough to let the data speak,
narrow enough to down-weight implausible extreme values.

## Step 4: Prior Predictive Check {#sec-part2-step4}

A prior predictive check asks: *if we sample parameter values from the priors
and generate data, do the simulated outcomes look like real-world health
spending?*
Public health expenditure as a share of GDP is typically between 0% and 30%.
Prior-predicted values far outside this range would signal that our priors
are miscalibrated.

```{r}
#| label: fig-prior-ppc-controls
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Prior predictive distribution: 2,000 parameter draws from the joint prior, each used to generate predicted health spending for all 120 observations. The shaded band marks the plausible 0–30% GDP range. Most prior mass falls inside — these priors are defensible."
set.seed(2026)
n_pp <- 2000

# For each of the 2000 draws, compute the linear predictor at every observation
# and add Gaussian noise with the drawn σ.
# rowwise() + mutate(y_sim = list(...)) creates one simulated dataset per draw.
prior_pred_controls <- tibble(
  b0 = rnorm(n_pp, 10, 5), # intercept draws — Normal(10, 5)
  b_wp = rnorm(n_pp, 0, 2), # women_parl slope draws — Normal(0, 2)
  b_gdp = rnorm(n_pp, 0, 2), # gdp_pc slope draws — Normal(0, 2)
  b_reg = rnorm(n_pp, 0, 2), # regime-type dummy draws — Normal(0, 2)
  sigma = rexp(n_pp, 1) # residual SD draws — Exponential(1)
) |>
  rowwise() |>
  mutate(
    y_sim = list(
      b0 +
        b_wp * df$women_parl +
        b_gdp * df$gdp_pc +
        b_reg * df$reg_type +
        rnorm(nrow(df), 0, sigma)
    )
  ) |>
  unnest(y_sim) |>
  ungroup()

prior_pred_controls |>
  ggplot(aes(x = y_sim)) +
  annotate(
    "rect",
    xmin = 0,
    xmax = 30,
    ymin = 0,
    ymax = Inf,
    fill = col_4,
    alpha = 0.12
  ) +
  geom_density(fill = col_1, alpha = 0.55, colour = "white", linewidth = 0.4) +
  geom_vline(
    xintercept = c(0, 30),
    colour = col_4,
    linetype = "dashed",
    linewidth = 0.9
  ) +
  annotate(
    "text",
    x = 15,
    y = Inf,
    vjust = 1.4,
    size = 3.2,
    colour = col_4,
    label = "Plausible empirical range: 0–30% GDP"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Predicted Health Expenditure (% GDP)",
    y = "Prior Predictive Density",
    title = "Prior Predictive Check: Health Spending Model",
    subtitle = "Before seeing data — does the prior generate plausible outcomes?"
  )
```

::: {.callout-note}
## Alternative implementation

```r
fit_prior_only <- brm(
  health_spend ~ women_parl + gdp_pc + reg_type,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2),  class = "b"),
    prior(exponential(1), class = "sigma")
  ),
  sample_prior = "only",
  chains = 2, iter = 1000, seed = 2026
)
pp_check(fit_prior_only, ndraws = 50)
```

:::

## Step 5: Fit the Model with `brms` {#sec-part2-step5}

The `brm()` call mirrors the formal model written in Step 3.
Note that we specify a separate `prior()` for each coefficient using `coef =`.
This is more precise than applying one flat prior to all slopes, and it
documents our assumptions in the code itself.

`brms` translates the model to Stan (C++), compiles it, and samples 4 parallel
Markov chains.
With `iter = 2000` and `warmup = 1000`, each chain contributes 1,000 post-warmup
draws, giving 4,000 total posterior samples of $(\alpha, \beta_X, \beta_Z, \sigma)$.

We fit **both** the bivariate (no controls) and additive multivariate models here so we
can compare their estimates and assess OVB in Step 7.

```{r}
#| label: fit-biv
#| output: false
# Bivariate model: health spending ~ women in parliament only
# This model omits GDP and regime type — will over-estimate the association (OVB)
fit_biv <- brm(
  health_spend ~ women_parl,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"), # baseline ~10% GDP
    prior(normal(0, 2), class = "b"), # weakly informative prior
    prior(exponential(1), class = "sigma") # residual SD < 5 almost certainly
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

::: {.callout-note}
## Alternative implementation

```r
# Inspect default prior structure before specifying your own
get_prior(
  health_spend ~ women_parl,
  data   = df,
  family = gaussian()
)
```

:::

```{r}
#| label: fit-multi
#| output: false
# Additive multivariate model: health spending ~ women + GDP + regime type
# Including GDP and regime type closes the back-door path and accounts for
# the intercept-level difference between system types
fit_multi <- brm(
  health_spend ~ women_parl + gdp_pc + reg_type,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2), class = "b", coef = "women_parl"), # weakly informative
    prior(normal(0, 2), class = "b", coef = "gdp_pc"), # weakly informative
    prior(normal(0, 2), class = "b", coef = "reg_type"), # parliamentary shift
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

::: {.callout-note}
## Alternative implementation

```r
# Confirm the priors actually applied to the fitted model
prior_summary(fit_multi)
```

:::

## Step 6a: MCMC Diagnostics — Traceplots {#sec-part2-step6a}

Before interpreting any results, we must confirm that MCMC has converged.
Three things to check:

1. **Rhat < 1.01** for every parameter: the chains have mixed — all four
   explored the same region of parameter space.
2. **Bulk_ESS and Tail_ESS > 1,000** for every parameter: we have enough
   effective samples to trust posterior summaries and credible intervals.
3. **Traceplots** look like "hairy caterpillars": each chain wanders randomly
   around the same range with no trends, drifts, or stuck periods.

If any diagnostic fails, stop and investigate before interpreting results.

```{r}
#| label: fig-trace-multi
#| echo: true
#| code-fold: true
#| fig-height: 3.6
#| dpi: 500
#| fig-cap: "Traceplots for the five model parameters. Each colour is one chain. The 'hairy caterpillar' pattern — chains overlapping and mixing rapidly — confirms convergence."
mcmc_trace(
  fit_multi,
  pars = c("b_Intercept", "b_women_parl", "b_gdp_pc", "b_reg_type", "sigma"),
  facet_args = list(ncol = 2)
) +
  scale_colour_manual(values = chain_cols) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90")
  ) +
  labs(
    title = "MCMC Traceplots: Health Spending Model",
    subtitle = "All chains must mix well before interpreting results"
  )
```

::: {.callout-note}
## Alternative implementation

```r
mcmc_rank_hist(
  fit_multi,
  pars = c("b_Intercept", "b_women_parl", "b_gdp_pc", "b_reg_type", "sigma")
) +
  labs(title = "Rank histograms: uniform bars confirm good mixing")
```

:::

## Step 6b: Posterior Predictive Check {#sec-part2-step6b}

A posterior predictive check (PPC) asks: *do the model's predictions look like the observed data?*
`pp_check()` draws 50 samples from the posterior predictive distribution
(light blue lines) and overlays them on the observed outcome distribution
(dark blue line).
A well-specified model should produce simulated outcome distributions that
closely envelope the observed one.

```{r}
#| label: fig-ppc-multi
#| echo: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Posterior predictive check for the health spending model. The dark line is the observed distribution; light blue lines are 50 draws from the posterior predictive. The Gaussian model fits well."
pp_check(fit_multi, ndraws = 4000) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Posterior Predictive Check: Health Spending Model",
    subtitle = "Light blue: model predictions  |  Dark: observed data"
  )
```

::: {.callout-note}
## Alternative implementation

```r
pp_check(fit_multi, type = "stat_2d", stat = c("mean", "sd")) +
  labs(title = "Joint PPC: posterior mean and SD vs observed")
```

:::

## Step 7: Interpret Results {#sec-part2-step7}

`summary()` returns posterior means (`Estimate`), posterior standard deviations
(`Est.Error`), and 95% credible intervals (`l-95% CI` / `u-95% CI`) for each
parameter.
The `Rhat` column confirms convergence (should be < 1.01 everywhere) and
`Bulk_ESS` / `Tail_ESS` confirm adequate effective sample sizes.

```{r}
#| label: controls-summary
#| echo: true
summary(fit_multi)
```

::: {.callout-note}
## Alternative implementation

```r
as_draws_df(fit_multi) |>
  select(b_women_parl, b_gdp_pc, b_reg_type, sigma) |>
  summarise(across(
    everything(),
    list(mean  = mean,
         sd    = sd,
         q2.5  = \(x) quantile(x, 0.025),
         q97.5 = \(x) quantile(x, 0.975))
  ))
```

:::

**What to look for**:

- `b_women_parl` posterior mean should be near the DGP average across both system types (approximately 0.08 + 0.10 × mean(`reg_type`) ≈ 0.135, since the additive model averages over the interaction).
  Compare this to the bivariate estimate — the difference is the OVB from omitting GDP.
- `b_gdp_pc` posterior mean should be near 0.20.
- `b_reg_type` posterior mean should be near 1.5 (parliamentary intercept shift).
- `sigma` should be near 1.5.

### OVB: Comparing Bivariate and Multivariate Estimates

We now compare the `women_parl` coefficient from both models.
The bivariate estimate (no GDP control) is inflated by OVB;
the multivariate estimate recovers a value close to the true DGP value of 0.12.

```{r}
#| label: fig-ovb-comparison
#| echo: true
#| code-fold: true
#| fig-height: 3.0
#| dpi: 500
#| fig-cap: "OVB visualised: posterior mean and 95% credible interval for the women_parl coefficient from the bivariate (orange) and multivariate (purple) models. The bivariate estimate is inflated by the GDP confounder; the multivariate model gives a better-controlled average slope across both regime types."
post_biv <- posterior_summary(fit_biv)["b_women_parl", ]
post_multi <- posterior_summary(fit_multi)["b_women_parl", ]

coef_df <- tibble(
  Model = factor(
    c("Bivariate\n(no controls)", "Multivariate\n(+ GDP + Regime Type)"),
    levels = c(
      "Bivariate\n(no controls)",
      "Multivariate\n(+ GDP + Regime Type)"
    )
  ),
  Estimate = c(post_biv["Estimate"], post_multi["Estimate"]),
  lo = c(post_biv["Q2.5"], post_multi["Q2.5"]),
  hi = c(post_biv["Q97.5"], post_multi["Q97.5"])
)

# DGP average slope: 0.08 + 0.10 * mean(reg_type) ≈ 0.135
dgp_avg <- 0.08 + 0.10 * mean(df$reg_type)

coef_df |>
  ggplot(aes(x = Model, y = Estimate)) +
  geom_hline(
    yintercept = dgp_avg,
    colour = "gray50",
    linetype = "dashed",
    linewidth = 0.9
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi, colour = Model),
    width = 0.12,
    linewidth = 1.1
  ) +
  geom_point(aes(colour = Model), size = 5) +
  scale_colour_manual(values = c(col_2, col_3), guide = "none") +
  annotate(
    "text",
    x = 1.52,
    y = dgp_avg + 0.005,
    size = 3.3,
    colour = "gray40",
    label = "DGP average slope (~0.135)",
    hjust = 0
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = NULL,
    y = expression(hat(beta) ~ "(women in parliament -> health spending)"),
    title = "OVB: Bivariate Inflates the Estimate",
    subtitle = "Adding GDP + regime type controls reduces the estimate toward the DGP average"
  ) +
  scale_y_continuous(limits = c(0, 0.40))
```

::: {.callout-note}
## Alternative implementation

```r
bind_rows(
  as_draws_df(fit_biv) |>
    transmute(b_women_parl, model = "Bivariate"),
  as_draws_df(fit_multi) |>
    transmute(b_women_parl, model = "Multivariate (+ GDP + Regime Type)")
) |>
  ggplot(aes(x = b_women_parl, fill = model)) +
  geom_density(alpha = 0.55) +
  scale_fill_manual(values = c(col_2, col_3)) +
  theme_bw() +
  labs(title = "Full posterior densities: OVB as distributional shift",
       x = expression(beta["women"]), fill = NULL)
```

:::

### Visualising the Partial Effect

Rather than just reporting the coefficient, we visualise the estimated
relationship.
We create a prediction grid that varies `women_parl` across its observed
range while holding `gdp_pc` fixed at its sample mean — this isolates the
partial effect of legislative representation.

`posterior_epred()` returns a matrix of posterior draws of $E[Y \mid X, Z]$
for each grid row.
Taking quantiles across draws gives us the posterior mean and 95% credible
interval for the partial effect at each value of `women_parl`.

```{r}
#| label: fig-controls-epred
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Partial effect of women's parliamentary share on health spending, holding GDP per capita at its sample mean. The ribbon is the 95% posterior credible interval."
# Grid: vary women_parl, hold gdp_pc at mean, regime type at modal value (1 = parliamentary)
grid_controls <- tibble(
  women_parl = seq(
    min(df$women_parl),
    max(df$women_parl),
    length.out = 80
  ),
  gdp_pc = mean(df$gdp_pc),
  reg_type = 1L # evaluate partial effect for parliamentary systems
)

# posterior_epred returns a draws × grid matrix of E[Y | X, Z, D] values
epred_mat <- posterior_epred(fit_multi, newdata = grid_controls)

grid_controls <- grid_controls |>
  mutate(
    lo = apply(epred_mat, 2, quantile, 0.025), # 2.5th percentile across draws
    mid = apply(epred_mat, 2, median), # median across draws
    hi = apply(epred_mat, 2, quantile, 0.975) # 97.5th percentile across draws
  )

df |>
  ggplot(aes(x = women_parl, y = health_spend)) +
  geom_ribbon(
    data = grid_controls,
    aes(x = women_parl, ymin = lo, ymax = hi, y = mid),
    inherit.aes = FALSE,
    fill = col_1,
    alpha = 0.25
  ) +
  geom_line(
    data = grid_controls,
    aes(x = women_parl, y = mid),
    inherit.aes = FALSE,
    colour = col_2,
    linewidth = 1.2
  ) +
  geom_point(colour = col_1, alpha = 0.45, size = 1.8) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Expenditure (% GDP)",
    title = "Partial Effect: Women in Parliament -> Health Spending",
    subtitle = paste0(
      "GDP per capita held at mean (",
      round(mean(df$gdp_pc), 1),
      " thousand USD)  |  Parliamentary systems  |  posterior_epred()"
    )
  )
```

::: {.callout-note}
## Alternative implementation

```r
# brms built-in: partial effect marginalised over other predictors
plot(conditional_effects(fit_multi, effects = "women_parl"))
```

:::

## Step 8: LOO-IC Model Comparison {#sec-part2-step8}

Bayesian leave-one-out cross-validation (LOO-IC) estimates how well each model
would predict *new* data — it penalises complexity implicitly.
We compare the bivariate model (no GDP control) against the multivariate model.

`add_criterion()` appends the LOO estimate to each fitted model.
`loo_compare()` ranks models by expected log predictive density (elpd):
higher elpd = better predictive accuracy.
When `|elpd_diff| > 2 × se_diff`, the evidence for the better model is
substantial.

```{r}
#| label: loo-controls-compare
#| echo: true
fit_multi <- add_criterion(fit_multi, "loo")
fit_biv <- add_criterion(fit_biv, "loo")
loo_compare(fit_biv, fit_multi)
```

::: {.callout-note}
## Alternative implementation

```r
# WAIC as alternative information criterion (same ranking expected)
fit_multi <- add_criterion(fit_multi, "waic")
fit_biv   <- add_criterion(fit_biv,   "waic")
loo_compare(fit_biv, fit_multi, criterion = "waic")
```

:::

The multivariate model should dominate: GDP and regime type both genuinely predict health spending,
so including them improves out-of-sample accuracy.

**Remaining limitations** to keep in mind:

1. GDP may not be the only confounder — country institutional quality, colonial
   history, and cultural factors might also drive both representation and spending.
2. Country-year observations within the same country are likely correlated —
   a violation of the independence assumption (A5) that Week 12 addresses with
   multilevel models.
3. Linearity is assumed: we should inspect residuals for non-linear patterns.


---

# Session Info and Execution Time {#sec-session}

```{r}
#| label: session-info
#| eval: true
sessionInfo()
```

```{r}
#| label: exec-time
#| eval: true
end_time <- Sys.time()
cat(
  "Total execution time:",
  round(difftime(end_time, start_time, units = "mins"), 1),
  "minutes\n"
)
```
