---
title: "Week 9 Lab: Multivariate Regression and Interactions"
subtitle: "Applications from the lecture — work through the code and check that it runs"
author: "Firstname Lastname"
date: last-modified
date-format: MMMM D, YYYY
format: html
toc: true
toc-depth: 3
code-fold: true
code-tools: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# Setup {#sec-setup}

Before running any analysis, we load the packages we need and define a few
shared colour constants that keep plots visually consistent.
`brms` fits Bayesian regression models via Stan; `bayesplot` produces
MCMC diagnostics; `patchwork` assembles multi-panel figures.

```{r}
#| label: setup
#| code-fold: false
start_time <- Sys.time() # used to measure execution time at the end

library(tidyverse) # data manipulation and ggplot2
library(brms) # Bayesian regression via Stan
library(bayesplot) # MCMC diagnostic plots
library(patchwork) # composing multi-panel figures

# Shared colour palette — reused in every plot so figures look consistent
col_1 <- "steelblue"
col_2 <- "#E07B39"
col_3 <- "#7B3F9E"
col_4 <- "goldenrod3"
chain_cols <- c(col_1, col_2, col_3, col_4) # one colour per MCMC chain
```

---

# Part II: Application I — Multivariate `brms` Model {#sec-part2}

## DAGs in Quarto: MermaidJS {#sec-dag-mermaid}

One quick way to sketch a DAG directly in a Quarto document — no R package required.
Use a fenced `{mermaid}` block with `graph LR` (left-to-right) or `graph TD` (top-down).

```{mermaid}
graph LR
  GDP["GDP per Capita"] --> WP["Women in Parliament (%)"]
  GDP --> HS["Health Spending (% GDP)"]
  WP --> HS
  RT["Regime Type"] --> HS
```

Nodes take labels in `["quoted text"]`; directed edges are `-->`.
This is enough for a homework submission or a quick theoretical sketch.

## DGP: Synthetic Data (Dataset 1) {#sec-part2-dgp}

This lab uses a **single synthetic dataset** throughout all four parts.
Working with synthetic data is pedagogically powerful: we know the true
parameter values, so we can verify whether our models recover them.
This is something you can almost never do with real data — in practice the DGP
is unobserved; we only see the outcomes, never the process that produced them.

The research question has two layers, addressed progressively:

1. **Parts I–II**: Is the share of women in parliament associated with higher public health expenditure, adjusting for GDP per capita and regime type?
2. **Parts III–IV**: Does this association differ between presidential and parliamentary systems?

The key complication in Part II is **confounding**: wealthier countries
simultaneously have more women in parliament *and* spend more on health.
Failing to account for GDP per capita will over-estimate the legislative
representation association.
In Part IV we add a second complication: the slope itself may differ across
regime types — this is **association heterogeneity**, addressed via an
interaction term.

```{r}
#| label: dgp
set.seed(2026) # reproducibility: same data every time
n <- 120 # 120 synthetic country-years

# ── Population-level parameters (the ground truth) ──────────────────────────
beta_0 <- 2 # intercept (presidential baseline)
beta_women_pres <- 0.08 # women_parl slope in presidential systems
beta_gdp <- 0.20 # GDP per capita partial association
beta_reg <- 1.5 # parliamentary intercept shift
beta_interact <- 0.10 # extra women_parl slope in parliamentary systems
sigma <- 1.5 # residual standard deviation

# ── Data-generating process ──────────────────────────────────────────────────
gdp_pc <- rnorm(n, mean = 18, sd = 7)
women_parl <- 15 + 0.9 * gdp_pc + rnorm(n, 0, 5)
reg_type <- rbinom(n, 1, prob = 0.55) # 1 = parliamentary, 0 = presidential

health_spend <- beta_0 +
  beta_gdp * gdp_pc +
  beta_women_pres * women_parl +
  beta_reg * reg_type +
  beta_interact * women_parl * reg_type + # steeper slope in parliamentary
  rnorm(n, 0, sigma)

df <- tibble(gdp_pc, women_parl, reg_type, health_spend) |>
  mutate(system = if_else(reg_type == 1, "Parliamentary", "Presidential"))
```

::: {.callout-note}
## Alternative implementation

```r
# OLS check: verify OVB direction without brms
m_biv  <- lm(health_spend ~ women_parl, data = df)
m_full <- lm(health_spend ~ women_parl + gdp_pc + reg_type, data = df)
round(coef(m_biv)["women_parl"] - coef(m_full)["women_parl"], 3)
```

:::

True women-in-parliament slopes:

| System | Slope |
|---|---|
| Presidential (`reg_type = 0`) | **0.08** |
| Parliamentary (`reg_type = 1`) | **0.08 + 0.10 = 0.18** |

Because GDP drives both variables, a bivariate model (no controls)
will produce an estimate well above either true value — that excess is omitted
variable bias (OVB).
The additive multivariate model (Parts I–II) estimates an *average* slope
across both system types; the interaction model (Parts III–IV) estimates
separate slopes.

## Step 1: Estimand {#sec-part2-step1}

Before writing any code, we state precisely what we are trying to estimate.
Drawing a DAG before analysis keeps this honest: the estimand follows from
the causal graph, not from what the model happens to produce.

Our DAG (presented in the lecture) shows GDP per capita as a **confounder** —
it has causal arrows into both women's parliamentary representation and health
spending.
A confounder opens a "back-door path" that conflates the association we care
about with the spurious GDP-driven correlation.
Conditioning on GDP per capita — including it as a predictor — closes that path.
Regime type (`reg_type`) enters as a **dummy variable** for the intercept shift
(parliamentary systems may have a different baseline level of spending) and
as the moderator in Part IV.

**Estimand (Part II)**: the partial association between women's parliamentary
representation and public health expenditure, adjusting for GDP per capita and
regime type — the average slope across both system types.

*Adjusting for GDP per capita and regime type shifts the estimate toward the true DGP average. Whether the adjustment set is sufficient is a substantive question the data alone cannot answer.*

## Step 2: Data Summary {#sec-part2-step2}

The scatter below shows the raw bivariate association between women's
parliamentary share and health spending.
There is a positive trend, but it conflates the legislative association with
the confounding from GDP.
We will visualise the OVB explicitly in Step 7, after fitting both models.

```{r}
#| label: fig-controls-raw
#| echo: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Raw bivariate scatter of women's parliamentary share against health expenditure. The positive trend conflates the legislative association with the GDP confounder."
df |>
  ggplot(aes(x = women_parl, y = health_spend)) +
  geom_point(colour = col_1, alpha = 0.55, size = 2) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Expenditure (% GDP)",
    title = "Raw Bivariate Association",
    subtitle = "Positive trend — but GDP confounds the relationship"
  )
```

::: {.callout-note}
## Alternative implementation

```r
df |>
  ggplot(aes(x = women_parl, y = health_spend, colour = gdp_pc)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_colour_gradient(low = col_1, high = col_2, name = "GDP pc") +
  theme_bw() +
  labs(title = "Confounding visible: GDP colours the point cloud",
       x = "Women in Parliament (%)", y = "Health Expenditure (% GDP)")
```

:::

## Step 3: Formal Model Specification {#sec-part2-step3}

We write out the full statistical model — likelihood and priors — before
touching `brms`.
This makes our assumptions explicit and contestable.

**Likelihood** (unchanged from Week 8 — still Gaussian):

$$
Y_i \sim \text{Normal}(\mu_i,\, \sigma)
$$

$$
\mu_i = \alpha + \beta_X X_i + \beta_Z Z_i + \beta_D D_i
$$

where $X$ = women in parliament (%), $Z$ = GDP per capita (thousands USD, PPP),
$D$ = regime type (1 = parliamentary, 0 = presidential).
The dummy $D$ gives parliamentary systems a different intercept baseline.

**Priors** (one per unknown parameter — all five must be justified):

$$
\alpha \sim \text{Normal}(10, 5) \qquad \beta_X \sim \text{Normal}(0, 2)
$$

$$
\beta_Z \sim \text{Normal}(0, 2) \qquad \beta_D \sim \text{Normal}(0, 2) \qquad \sigma \sim \text{Exponential}(1)
$$

The prior on $\beta_X$ says: *before seeing data, a 1 pp increase in women's
representation is associated with a change of somewhere between −6 and +6 pp
in health spending (the Normal(0, 2) range at ±3σ is [−6, +6]).*
The prior on $\beta_D$ says: *parliamentary systems may spend anywhere from about 6 pp
less to 6 pp more than presidential systems, holding other predictors constant —
a weakly informative, non-dogmatic prior with no assumption about which system spends more.*
Both are defensible weakly informative priors: broad enough to let the data speak,
narrow enough to down-weight implausible extreme values.

## Step 4: Prior Predictive Check {#sec-part2-step4}

A prior predictive check asks: *if we sample parameter values from the priors
and generate data, do the simulated outcomes look like real-world health
spending?*
Public health expenditure as a share of GDP is typically between 0% and 30%.
Prior-predicted values far outside this range would signal that our priors
are miscalibrated.

```{r}
#| label: fig-prior-ppc-controls
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Prior predictive distribution: 2,000 parameter draws from the joint prior, each used to generate predicted health spending for all 120 observations. The shaded band marks the plausible 0–30% GDP range. Most prior mass falls inside — these priors are defensible."
set.seed(2026)
n_pp <- 2000

# For each of the 2000 draws, compute the linear predictor at every observation
# and add Gaussian noise with the drawn σ.
# rowwise() + mutate(y_sim = list(...)) creates one simulated dataset per draw.
prior_pred_controls <- tibble(
  b0 = rnorm(n_pp, 10, 5), # intercept draws — Normal(10, 5)
  b_wp = rnorm(n_pp, 0, 2), # women_parl slope draws — Normal(0, 2)
  b_gdp = rnorm(n_pp, 0, 2), # gdp_pc slope draws — Normal(0, 2)
  b_reg = rnorm(n_pp, 0, 2), # regime-type dummy draws — Normal(0, 2)
  sigma = rexp(n_pp, 1) # residual SD draws — Exponential(1)
) |>
  rowwise() |>
  mutate(
    y_sim = list(
      b0 +
        b_wp * df$women_parl +
        b_gdp * df$gdp_pc +
        b_reg * df$reg_type +
        rnorm(nrow(df), 0, sigma)
    )
  ) |>
  unnest(y_sim) |>
  ungroup()

prior_pred_controls |>
  ggplot(aes(x = y_sim)) +
  annotate(
    "rect",
    xmin = 0,
    xmax = 30,
    ymin = 0,
    ymax = Inf,
    fill = col_4,
    alpha = 0.12
  ) +
  geom_density(fill = col_1, alpha = 0.55, colour = "white", linewidth = 0.4) +
  geom_vline(
    xintercept = c(0, 30),
    colour = col_4,
    linetype = "dashed",
    linewidth = 0.9
  ) +
  annotate(
    "text",
    x = 15,
    y = Inf,
    vjust = 1.4,
    size = 3.2,
    colour = col_4,
    label = "Plausible empirical range: 0–30% GDP"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Predicted Health Expenditure (% GDP)",
    y = "Prior Predictive Density",
    title = "Prior Predictive Check: Health Spending Model",
    subtitle = "Before seeing data — does the prior generate plausible outcomes?"
  )
```

::: {.callout-note}
## Alternative implementation

```r
fit_prior_only <- brm(
  health_spend ~ women_parl + gdp_pc + reg_type,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2),  class = "b"),
    prior(exponential(1), class = "sigma")
  ),
  sample_prior = "only",
  chains = 2, iter = 1000, seed = 2026
)
pp_check(fit_prior_only, ndraws = 50)
```

:::

## Step 5: Fit the Model with `brms` {#sec-part2-step5}

The `brm()` call mirrors the formal model written in Step 3.
Note that we specify a separate `prior()` for each coefficient using `coef =`.
This is more precise than applying one flat prior to all slopes, and it
documents our assumptions in the code itself.

`brms` translates the model to Stan (C++), compiles it, and samples 4 parallel
Markov chains.
With `iter = 2000` and `warmup = 1000`, each chain contributes 1,000 post-warmup
draws, giving 4,000 total posterior samples of $(\alpha, \beta_X, \beta_Z, \sigma)$.

We fit **both** the bivariate (no controls) and additive multivariate models here so we
can compare their estimates and assess OVB in Step 7.

```{r}
#| label: fit-biv
#| output: false
# Bivariate model: health spending ~ women in parliament only
# This model omits GDP and regime type — will over-estimate the association (OVB)
fit_biv <- brm(
  health_spend ~ women_parl,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"), # baseline ~10% GDP
    prior(normal(0, 2), class = "b"), # weakly informative prior
    prior(exponential(1), class = "sigma") # residual SD < 5 almost certainly
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

::: {.callout-note}
## Alternative implementation

```r
# Inspect default prior structure before specifying your own
get_prior(
  health_spend ~ women_parl,
  data   = df,
  family = gaussian()
)
```

:::

```{r}
#| label: fit-multi
#| output: false
# Additive multivariate model: health spending ~ women + GDP + regime type
# Including GDP and regime type closes the back-door path and accounts for
# the intercept-level difference between system types
fit_multi <- brm(
  health_spend ~ women_parl + gdp_pc + reg_type,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2), class = "b", coef = "women_parl"), # weakly informative
    prior(normal(0, 2), class = "b", coef = "gdp_pc"), # weakly informative
    prior(normal(0, 2), class = "b", coef = "reg_type"), # parliamentary shift
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

::: {.callout-note}
## Alternative implementation

```r
# Confirm the priors actually applied to the fitted model
prior_summary(fit_multi)
```

:::

## Step 6a: MCMC Diagnostics — Traceplots {#sec-part2-step6a}

Before interpreting any results, we must confirm that MCMC has converged.
Three things to check:

1. **Rhat < 1.01** for every parameter: the chains have mixed — all four
   explored the same region of parameter space.
2. **Bulk_ESS and Tail_ESS > 1,000** for every parameter: we have enough
   effective samples to trust posterior summaries and credible intervals.
3. **Traceplots** look like "hairy caterpillars": each chain wanders randomly
   around the same range with no trends, drifts, or stuck periods.

If any diagnostic fails, stop and investigate before interpreting results.

```{r}
#| label: fig-trace-multi
#| echo: true
#| code-fold: true
#| fig-height: 3.6
#| dpi: 500
#| fig-cap: "Traceplots for the five model parameters. Each colour is one chain. The 'hairy caterpillar' pattern — chains overlapping and mixing rapidly — confirms convergence."
mcmc_trace(
  fit_multi,
  pars = c("b_Intercept", "b_women_parl", "b_gdp_pc", "b_reg_type", "sigma"),
  facet_args = list(ncol = 2)
) +
  scale_colour_manual(values = chain_cols) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90")
  ) +
  labs(
    title = "MCMC Traceplots: Health Spending Model",
    subtitle = "All chains must mix well before interpreting results"
  )
```

::: {.callout-note}
## Alternative implementation

```r
mcmc_rank_hist(
  fit_multi,
  pars = c("b_Intercept", "b_women_parl", "b_gdp_pc", "b_reg_type", "sigma")
) +
  labs(title = "Rank histograms: uniform bars confirm good mixing")
```

:::

## Step 6b: Posterior Predictive Check {#sec-part2-step6b}

A posterior predictive check (PPC) asks: *do the model's predictions look like the observed data?*
`pp_check()` draws 50 samples from the posterior predictive distribution
(light blue lines) and overlays them on the observed outcome distribution
(dark blue line).
A well-specified model should produce simulated outcome distributions that
closely envelope the observed one.

```{r}
#| label: fig-ppc-multi
#| echo: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Posterior predictive check for the health spending model. The dark line is the observed distribution; light blue lines are 50 draws from the posterior predictive. The Gaussian model fits well."
pp_check(fit_multi, ndraws = 4000) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Posterior Predictive Check: Health Spending Model",
    subtitle = "Light blue: model predictions  |  Dark: observed data"
  )
```

::: {.callout-note}
## Alternative implementation

```r
pp_check(fit_multi, type = "stat_2d", stat = c("mean", "sd")) +
  labs(title = "Joint PPC: posterior mean and SD vs observed")
```

:::

## Step 7: Interpret Results {#sec-part2-step7}

`summary()` returns posterior means (`Estimate`), posterior standard deviations
(`Est.Error`), and 95% credible intervals (`l-95% CI` / `u-95% CI`) for each
parameter.
The `Rhat` column confirms convergence (should be < 1.01 everywhere) and
`Bulk_ESS` / `Tail_ESS` confirm adequate effective sample sizes.

```{r}
#| label: controls-summary
#| echo: true
summary(fit_multi)
```

::: {.callout-note}
## Alternative implementation

```r
as_draws_df(fit_multi) |>
  select(b_women_parl, b_gdp_pc, b_reg_type, sigma) |>
  summarise(across(
    everything(),
    list(mean  = mean,
         sd    = sd,
         q2.5  = \(x) quantile(x, 0.025),
         q97.5 = \(x) quantile(x, 0.975))
  ))
```

:::

**What to look for**:

- `b_women_parl` posterior mean should be near the DGP average across both system types (approximately 0.08 + 0.10 × mean(`reg_type`) ≈ 0.135, since the additive model averages over the interaction).
  Compare this to the bivariate estimate — the difference is the OVB from omitting GDP.
- `b_gdp_pc` posterior mean should be near 0.20.
- `b_reg_type` posterior mean should be near 1.5 (parliamentary intercept shift).
- `sigma` should be near 1.5.

### OVB: Comparing Bivariate and Multivariate Estimates

We now compare the `women_parl` coefficient from both models.
The bivariate estimate (no GDP control) is inflated by OVB;
the multivariate estimate recovers a value close to the true DGP value of 0.12.

```{r}
#| label: fig-ovb-comparison
#| echo: true
#| code-fold: true
#| fig-height: 3.0
#| dpi: 500
#| fig-cap: "OVB visualised: posterior mean and 95% credible interval for the women_parl coefficient from the bivariate (orange) and multivariate (purple) models. The bivariate estimate is inflated by the GDP confounder; the multivariate model gives a better-controlled average slope across both regime types."
post_biv <- posterior_summary(fit_biv)["b_women_parl", ]
post_multi <- posterior_summary(fit_multi)["b_women_parl", ]

coef_df <- tibble(
  Model = factor(
    c("Bivariate\n(no controls)", "Multivariate\n(+ GDP + Regime Type)"),
    levels = c(
      "Bivariate\n(no controls)",
      "Multivariate\n(+ GDP + Regime Type)"
    )
  ),
  Estimate = c(post_biv["Estimate"], post_multi["Estimate"]),
  lo = c(post_biv["Q2.5"], post_multi["Q2.5"]),
  hi = c(post_biv["Q97.5"], post_multi["Q97.5"])
)

# DGP average slope: 0.08 + 0.10 * mean(reg_type) ≈ 0.135
dgp_avg <- 0.08 + 0.10 * mean(df$reg_type)

coef_df |>
  ggplot(aes(x = Model, y = Estimate)) +
  geom_hline(
    yintercept = dgp_avg,
    colour = "gray50",
    linetype = "dashed",
    linewidth = 0.9
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi, colour = Model),
    width = 0.12,
    linewidth = 1.1
  ) +
  geom_point(aes(colour = Model), size = 5) +
  scale_colour_manual(values = c(col_2, col_3), guide = "none") +
  annotate(
    "text",
    x = 1.52,
    y = dgp_avg + 0.005,
    size = 3.3,
    colour = "gray40",
    label = "DGP average slope (~0.135)",
    hjust = 0
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = NULL,
    y = expression(hat(beta) ~ "(women in parliament -> health spending)"),
    title = "OVB: Bivariate Inflates the Estimate",
    subtitle = "Adding GDP + regime type controls reduces the estimate toward the DGP average"
  ) +
  scale_y_continuous(limits = c(0, 0.40))
```

::: {.callout-note}
## Alternative implementation

```r
bind_rows(
  as_draws_df(fit_biv) |>
    transmute(b_women_parl, model = "Bivariate"),
  as_draws_df(fit_multi) |>
    transmute(b_women_parl, model = "Multivariate (+ GDP + Regime Type)")
) |>
  ggplot(aes(x = b_women_parl, fill = model)) +
  geom_density(alpha = 0.55) +
  scale_fill_manual(values = c(col_2, col_3)) +
  theme_bw() +
  labs(title = "Full posterior densities: OVB as distributional shift",
       x = expression(beta["women"]), fill = NULL)
```

:::

### Visualising the Partial Effect

Rather than just reporting the coefficient, we visualise the estimated
relationship.
We create a prediction grid that varies `women_parl` across its observed
range while holding `gdp_pc` fixed at its sample mean — this isolates the
partial effect of legislative representation.

`posterior_epred()` returns a matrix of posterior draws of $E[Y \mid X, Z]$
for each grid row.
Taking quantiles across draws gives us the posterior mean and 95% credible
interval for the partial effect at each value of `women_parl`.

```{r}
#| label: fig-controls-epred
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Partial effect of women's parliamentary share on health spending, holding GDP per capita at its sample mean. The ribbon is the 95% posterior credible interval."
# Grid: vary women_parl, hold gdp_pc at mean, regime type at modal value (1 = parliamentary)
grid_controls <- tibble(
  women_parl = seq(
    min(df$women_parl),
    max(df$women_parl),
    length.out = 80
  ),
  gdp_pc = mean(df$gdp_pc),
  reg_type = 1L # evaluate partial effect for parliamentary systems
)

# posterior_epred returns a draws × grid matrix of E[Y | X, Z, D] values
epred_mat <- posterior_epred(fit_multi, newdata = grid_controls)

grid_controls <- grid_controls |>
  mutate(
    lo = apply(epred_mat, 2, quantile, 0.025), # 2.5th percentile across draws
    mid = apply(epred_mat, 2, median), # median across draws
    hi = apply(epred_mat, 2, quantile, 0.975) # 97.5th percentile across draws
  )

df |>
  ggplot(aes(x = women_parl, y = health_spend)) +
  geom_ribbon(
    data = grid_controls,
    aes(x = women_parl, ymin = lo, ymax = hi, y = mid),
    inherit.aes = FALSE,
    fill = col_1,
    alpha = 0.25
  ) +
  geom_line(
    data = grid_controls,
    aes(x = women_parl, y = mid),
    inherit.aes = FALSE,
    colour = col_2,
    linewidth = 1.2
  ) +
  geom_point(colour = col_1, alpha = 0.45, size = 1.8) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Expenditure (% GDP)",
    title = "Partial Effect: Women in Parliament -> Health Spending",
    subtitle = paste0(
      "GDP per capita held at mean (",
      round(mean(df$gdp_pc), 1),
      " thousand USD)  |  Parliamentary systems  |  posterior_epred()"
    )
  )
```

::: {.callout-note}
## Alternative implementation

```r
# brms built-in: partial effect marginalised over other predictors
plot(conditional_effects(fit_multi, effects = "women_parl"))
```

:::

## Step 8: LOO-IC Model Comparison {#sec-part2-step8}

Bayesian leave-one-out cross-validation (LOO-IC) estimates how well each model
would predict *new* data — it penalises complexity implicitly.
We compare the bivariate model (no GDP control) against the multivariate model.

`add_criterion()` appends the LOO estimate to each fitted model.
`loo_compare()` ranks models by expected log predictive density (elpd):
higher elpd = better predictive accuracy.
When `|elpd_diff| > 2 × se_diff`, the evidence for the better model is
substantial.

```{r}
#| label: loo-controls-compare
#| echo: true
fit_multi <- add_criterion(fit_multi, "loo")
fit_biv <- add_criterion(fit_biv, "loo")
loo_compare(fit_biv, fit_multi)
```

::: {.callout-note}
## Alternative implementation

```r
# WAIC as alternative information criterion (same ranking expected)
fit_multi <- add_criterion(fit_multi, "waic")
fit_biv   <- add_criterion(fit_biv,   "waic")
loo_compare(fit_biv, fit_multi, criterion = "waic")
```

:::

The multivariate model should dominate: GDP and regime type both genuinely predict health spending,
so including them improves out-of-sample accuracy.

**Remaining limitations** to keep in mind:

1. GDP may not be the only confounder — country institutional quality, colonial
   history, and cultural factors might also drive both representation and spending.
2. Country-year observations within the same country are likely correlated —
   a violation of the independence assumption (A5) that Week 12 addresses with
   multilevel models.
3. Linearity is assumed: we should inspect residuals for non-linear patterns.

---

# Part IV: Application II — Interaction `brms` Model {#sec-part4}

## The Research Question Revisited {#sec-part4-intro}

Parts I–II established that women's parliamentary representation is associated
with higher health spending, even after controlling for GDP per capita and regime
type as an additive intercept shift.
But the additive model assumes the **slope** of the women-in-parliament
association is the same for both presidential and parliamentary systems.

Part IV relaxes this assumption:

> **Does the association between female legislator share and health spending vary by regime type?**

**Legislative composition theory**: parliamentary systems concentrate budget
authority in the legislature — strong party discipline, parliamentary
sovereignty, budget processes rooted in the legislature.
Presidential systems divide power across branches, potentially weakening the
translation of legislative preferences into policy.
This predicts a **steeper slope** for women in parliament in parliamentary systems.

We use the **same `df` dataset** as Parts I–II.
The true DGP already encodes this interaction:

| System | Women-in-parliament slope |
|---|---|
| Presidential (`reg_type = 0`) | **0.08** |
| Parliamentary (`reg_type = 1`) | **0.08 + 0.10 = 0.18** |

The interaction coefficient **$\beta_3 = 0.10$** captures *how much steeper*
the parliamentary slope is relative to the presidential slope.

## Step 1: DAG and Estimand {#sec-part4-step1}

The DAG (introduced in the lecture) shows regime type having a **direct arrow**
into health spending (intercept shift, captured in Parts I–II) and a **moderation
role** on the women-in-parliament → health-spending edge (the interaction, the
focus of Part IV).

Because we already control for GDP per capita (the confounder), no additional
adjustment is needed.
The interaction term itself is the new quantity of interest.

**Estimand 1**: $\partial\mu/\partial X \mid Z=0$ — women-in-parliament slope
in *presidential* systems.

**Estimand 2**: $\partial\mu/\partial X \mid Z=1$ — women-in-parliament slope
in *parliamentary* systems.

**Estimand 3**: $\beta_3$ — the slope difference; how much more does women's
representation associate with health spending in parliamentary vs. presidential
systems?

## Step 2: Data Summary {#sec-part4-step2}

We start with the marginal distribution of health spending by system type,
then examine the joint scatter with separate OLS lines per group and a
tercile-based boxplot comparison.

```{r}
#| label: fig-outcome-dist-interact
#| echo: true
#| fig-height: 3.0
#| dpi: 500
#| fig-cap: "Marginal distribution of health spending by regime type. Parliamentary systems show a higher average — the intercept shift β₂ captured in the additive model (Parts I–II)."
df |>
  ggplot(aes(x = health_spend, fill = system, colour = system)) +
  geom_density(alpha = 0.45, linewidth = 0.8) +
  scale_fill_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2)
  ) +
  scale_colour_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2)
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Health Spending (% GDP)",
    y = "Density",
    title = "Distribution of Health Spending by Regime Type",
    subtitle = "Parliamentary systems spend more on average",
    fill = NULL,
    colour = NULL
  )
```

::: {.callout-note}
## Alternative implementation

```r
df |>
  ggplot(aes(x = system, y = health_spend, fill = system)) +
  geom_violin(alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.35, size = 1.2) +
  scale_fill_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2), guide = "none"
  ) +
  theme_bw() +
  labs(title = "Health spending by regime type — violin view",
       x = NULL, y = "Health Spending (% GDP)")
```

:::

The scatter below puts women in parliament on the x-axis and draws separate OLS
lines for each system type.
If the lines have visibly different slopes, that is preliminary evidence for
an interaction.
The boxplot facets women in parliament into terciles and shows whether the gap
between systems widens with more female legislators — the visual signature of
$\beta_3 > 0$.

```{r}
#| label: fig-interact-summary
#| echo: true
#| code-fold: true
#| fig-height: 3.6
#| dpi: 500
#| fig-cap: "Left: scatter with separate OLS lines per system — slopes appear different. Right: health spending by women-in-parliament tercile and system type — if the gap widens with more women in parliament, that is the visual signature of a positive interaction."
p_scatter_int <- df |>
  ggplot(aes(x = women_parl, y = health_spend, colour = system)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(
    aes(group = system),
    method = "lm",
    se = FALSE,
    linewidth = 1.2
  ) +
  scale_colour_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2),
    name = "System"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Spending (% GDP)",
    title = "Separate Slopes by Regime Type",
    subtitle = "OLS line per system"
  )

df_interact_plot <- df |>
  mutate(
    wp_tercile = cut(
      women_parl,
      breaks = quantile(women_parl, c(0, 1 / 3, 2 / 3, 1)),
      labels = c("Low\n(<25%)", "Mid\n(~33%)", "High\n(>40%)"),
      include.lowest = TRUE
    )
  )

p_box_int <- df_interact_plot |>
  ggplot(aes(x = wp_tercile, y = health_spend, fill = system)) +
  geom_boxplot(
    alpha = 0.7,
    position = position_dodge(width = 0.75),
    width = 0.6
  ) +
  scale_fill_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2),
    name = "System"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Women in Parliament Tercile",
    y = "Health Spending (% GDP)",
    title = "Gap Across Systems by Women-in-Parliament Level",
    subtitle = "Widening gap = signature of a positive interaction"
  )

p_scatter_int | p_box_int
```

::: {.callout-note}
## Alternative implementation

```r
df |>
  mutate(tercile = cut(women_parl,
    breaks = quantile(women_parl, c(0, 1/3, 2/3, 1)),
    labels = c("Low WP", "Mid WP", "High WP"),
    include.lowest = TRUE)) |>
  ggplot(aes(x = tercile, y = health_spend, fill = system)) +
  geom_boxplot(position = position_dodge(0.75), alpha = 0.7) +
  scale_fill_manual(values = c("Parliamentary" = col_1, "Presidential" = col_2)) +
  theme_bw() +
  labs(title = "Spending gap across systems — tercile view",
       x = "Women in Parliament Tercile", y = "Health Spending (% GDP)", fill = NULL)
```

:::

## Step 3: Formal Model Specification {#sec-part4-step3}

The interaction model augments the additive model with a product term:

$$
Y_i \sim \text{Normal}(\mu_i,\, \sigma)
$$

$$
\mu_i = \alpha + \beta_1 X_i + \beta_2 D_i + \beta_3 (X_i \cdot D_i) + \beta_Z Z_i
$$

where $X$ = women in parliament (%), $D$ = regime type (1 = parliamentary),
$Z$ = GDP per capita.

The two conditional marginal effects are now *functions* of the moderator:

$$
\frac{\partial\mu}{\partial X} = \beta_1 + \beta_3 D \qquad
\frac{\partial\mu}{\partial D} = \beta_2 + \beta_3 X
$$

**Priors**:

$$
\alpha \sim \text{Normal}(10, 5) \qquad
\beta_1, \beta_2, \beta_Z \sim \text{Normal}(0, 2) \qquad
\beta_3 \sim \text{Normal}(0, 2) \qquad
\sigma \sim \text{Exponential}(1)
$$

The prior for $\beta_3$ is centred at zero — our prior expectation is *no moderation*.
The data will update this.
This is appropriately humble: we should not encode strong prior beliefs about
the direction of an interaction before seeing the evidence.

We fit two competing models:

- **`fit_multi`** (no interaction): assumes the women-in-parliament slope is the
  same for all system types — a testable restriction (already fitted in Part II).
- **`fit_interact`** (interaction): relaxes this restriction.
  If `fit_interact` fits substantially better via LOO-IC, we have evidence that the
  restriction is false.

## Step 4: Prior Predictive Check {#sec-part4-step4}

For the interaction model, we draw regression lines from the joint prior —
one bundle per regime type.
The check asks: *does the prior allow a wide enough range of slope differences,
including no moderation (parallel lines), strong moderation (strongly diverging
lines), and implausible configurations?*

```{r}
#| label: fig-prior-ppc-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.4
#| dpi: 500
#| fig-cap: "Prior predictive spaghetti: 150 regression lines drawn from the joint prior, separately for parliamentary (blue) and presidential (orange) systems. The prior is wide but not implausible — it allows no interaction, moderate interaction, and strong interaction."
set.seed(2026)
n_pp2 <- 150
wp_seq <- seq(10, 55, length.out = 60)
gdp_mean <- mean(df$gdp_pc)

prior_lines_int <- tibble(
  sim = 1:n_pp2,
  b0 = rnorm(n_pp2, 10, 5), # intercept draws — Normal(10, 5)
  b_wp = rnorm(n_pp2, 0, 2), # women_parl slope draws — Normal(0, 2)
  b_reg = rnorm(n_pp2, 0, 2), # regime-type main effect draws — Normal(0, 2)
  b_int = rnorm(n_pp2, 0, 2), # interaction term draws — Normal(0, 2)
  b_gdp = rnorm(n_pp2, 0, 2) # GDP slope draws — Normal(0, 2)
) |>
  expand_grid(women_parl = wp_seq, reg_type = c(0L, 1L)) |>
  mutate(
    y_hat = b0 +
      b_wp * women_parl +
      b_reg * reg_type +
      b_int * women_parl * reg_type + # slope shifts by b_int in parliamentary
      b_gdp * gdp_mean,
    system = if_else(reg_type == 1, "Parliamentary", "Presidential")
  )

prior_lines_int |>
  ggplot(aes(
    x = women_parl,
    y = y_hat,
    group = interaction(sim, system),
    colour = system
  )) +
  geom_line(alpha = 0.12, linewidth = 0.5) +
  scale_colour_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2),
    name = "System"
  ) +
  coord_cartesian(ylim = c(-5, 30)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Women in Parliament (%)",
    y = "Predicted Health Spending (% GDP)",
    title = "Prior Predictive Spaghetti: Interaction Model",
    subtitle = "150 lines per system — prior is non-dogmatic about moderation"
  )
```

## Step 5: Fit Both Models {#sec-part4-step5}

We use the `*` operator in the `brm()` formula.
`women_parl * reg_type` is shorthand for `women_parl + reg_type + women_parl:reg_type`.
Always use `*`, not just the colon form (`:`).
The colon-only form omits the main effects, which forces $\beta_1 = 0$ and
$\beta_2 = 0$ — usually a false and arbitrary constraint.

`fit_multi` (the additive model, no interaction) was already fitted in Part II.
We fit `fit_interact` here.

```{r}
#| label: fit-interact
#| echo: true
#| output: false
# Interaction model — allows the women_parl slope to differ by regime type
# * expands to: women_parl + reg_type + women_parl:reg_type
fit_interact <- brm(
  health_spend ~ women_parl * reg_type + gdp_pc,
  data = df,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2), class = "b", coef = "women_parl"), # presidential slope
    prior(normal(0, 2), class = "b", coef = "reg_type"), # parliamentary shift
    prior(normal(0, 2), class = "b", coef = "gdp_pc"), # GDP partial effect
    prior(normal(0, 2), class = "b", coef = "women_parl:reg_type"), # slope difference
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

::: {.callout-note}
## Alternative implementation

```r
# Confirm priors actually applied to the interaction model
prior_summary(fit_interact)
```

:::

## Step 6: Diagnostics and Posterior Predictive Check {#sec-part4-step6}

We run the same convergence checks as before.
The traceplots and PPC are shown side-by-side for the interaction model.

```{r}
#| label: fig-diag-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Left: traceplots for the five parameters of the interaction model — all chains mix well. Right: posterior predictive check — model predictions envelope the observed distribution."
p_trace_int <- mcmc_trace(
  fit_interact,
  pars = c(
    "b_Intercept",
    "b_women_parl",
    "b_reg_type",
    "b_women_parl:reg_type",
    "b_gdp_pc"
  ),
  facet_args = list(ncol = 2)
) +
  scale_colour_manual(values = chain_cols) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90"),
    legend.position = "none"
  ) +
  labs(title = "Traceplots: Interaction Model")

p_ppc_int <- pp_check(fit_interact, ndraws = 50) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "Posterior Predictive Check")

p_trace_int | p_ppc_int
```

::: {.callout-note}
## Alternative implementation

```r
mcmc_rank_hist(
  fit_interact,
  pars = c("b_Intercept", "b_women_parl",
           "b_reg_type", "b_women_parl:reg_type", "b_gdp_pc")
) +
  labs(title = "Rank histograms: interaction model convergence check")
```

:::

## Step 7: Interpreting Results and Posterior Marginal Effects {#sec-part4-step7}

Inspect the summary output.
The key quantity is `b_women_parl:reg_type` — the interaction coefficient $\beta_3$.
Its posterior mean should be near 0.10 (the true DGP value).

```{r}
#| label: interact-summary
#| echo: true
summary(fit_interact)
```

::: {.callout-note}
## Alternative implementation

```r
as_draws_df(fit_interact) |>
  mutate(
    slope_presidential  = b_women_parl,
    slope_parliamentary = b_women_parl + `b_women_parl:reg_type`
  ) |>
  summarise(across(
    c(slope_presidential, slope_parliamentary),
    list(mean  = mean,
         q2.5  = \(x) quantile(x, 0.025),
         q97.5 = \(x) quantile(x, 0.975))
  ))
```

:::

### Posterior Line Clouds

For each MCMC draw we compute the intercept and slope for each system type
by simple arithmetic and draw the resulting line.
The bundle of parliamentary lines (blue) should be steeper than the presidential
bundle (orange).

```{r}
#| label: fig-posterior-lines-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Posterior regression line clouds: 200 MCMC draws per system type. The blue bundle (parliamentary) is visibly steeper — posterior evidence for β₃ > 0."
set.seed(2026)
post_draws_int <- as_draws_df(fit_interact) |> slice_sample(n = 200)
gdp_mean <- mean(df$gdp_pc)

df |>
  ggplot(aes(x = women_parl, y = health_spend, colour = system)) +
  # Presidential lines: intercept = b_Intercept + b_gdp_pc * gdp_mean,
  #                     slope     = b_women_parl
  geom_abline(
    data = post_draws_int,
    aes(
      intercept = b_Intercept + b_gdp_pc * gdp_mean,
      slope = b_women_parl
    ),
    colour = col_2,
    alpha = 0.05,
    linewidth = 0.4
  ) +
  # Parliamentary lines: intercept = b_Intercept + b_reg_type + b_gdp_pc * gdp_mean,
  #                      slope     = b_women_parl + b_women_parl:reg_type
  geom_abline(
    data = post_draws_int,
    aes(
      intercept = b_Intercept + b_reg_type + b_gdp_pc * gdp_mean,
      slope = b_women_parl + `b_women_parl:reg_type`
    ),
    colour = col_1,
    alpha = 0.05,
    linewidth = 0.4
  ) +
  geom_point(alpha = 0.5, size = 1.8) +
  scale_colour_manual(
    values = c("Parliamentary" = col_1, "Presidential" = col_2),
    name = "System"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Spending (% GDP)",
    title = "Posterior Line Clouds by Regime Type",
    subtitle = "200 MCMC draws per group — blue bundle (parliamentary) is steeper"
  )
```

::: {.callout-note}
## Alternative implementation

```r
# brms built-in: automatic interaction plot
plot(conditional_effects(
  fit_interact,
  effects = "women_parl:reg_type"
))
```

:::

### Posterior Conditional Marginal Effects

The conditional marginal effect of women in parliament for a given system type
is a function of $\beta_1$ and $\beta_3$:

$$
\text{ME}_{\text{presidential}}  = \beta_1 \qquad
\text{ME}_{\text{parliamentary}} = \beta_1 + \beta_3
$$

Because we have full posterior draws of both $\beta_1$ and $\beta_3$, we compute
these by arithmetic — no approximation or delta method needed.

```{r}
#| label: fig-marginal-effects-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.4
#| dpi: 500
#| fig-cap: "Posterior distributions of the conditional marginal effect of women in parliament on health spending by system type. The parliamentary slope (~0.18) and presidential slope (~0.08) should be well separated — posterior evidence for β₃ > 0."
post_all_int <- as_draws_df(fit_interact)

me_df <- post_all_int |>
  transmute(
    `Presidential` = b_women_parl, # ME at D = 0
    `Parliamentary` = b_women_parl + `b_women_parl:reg_type` # ME at D = 1
  ) |>
  pivot_longer(
    everything(),
    names_to = "system",
    values_to = "marginal_effect"
  ) |>
  mutate(
    system = factor(system, levels = c("Presidential", "Parliamentary"))
  )

# P(β₃ > 0 | data): direct posterior probability — not a p-value
prob_b3_pos <- mean(post_all_int$`b_women_parl:reg_type` > 0)

me_df |>
  ggplot(aes(x = marginal_effect, fill = system, colour = system)) +
  geom_density(alpha = 0.55, linewidth = 0.8) +
  geom_vline(
    xintercept = 0,
    colour = "gray40",
    linetype = "dotted",
    linewidth = 0.9
  ) +
  scale_fill_manual(
    values = c("Presidential" = col_2, "Parliamentary" = col_1),
    name = NULL
  ) +
  scale_colour_manual(
    values = c("Presidential" = col_2, "Parliamentary" = col_1),
    name = NULL
  ) +
  annotate(
    "text",
    x = max(me_df$marginal_effect) * 0.7,
    y = Inf,
    vjust = 1.5,
    size = 3.5,
    colour = col_3,
    label = paste0("P(\u03b2\u2083 > 0 | data) = ", round(prob_b3_pos, 3))
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = expression(partialdiff * mu / partialdiff ~ "Women in Parliament (%)"),
    y = "Posterior Density",
    title = "Posterior Marginal Effects: Women in Parliament -> Health Spending",
    subtitle = "Parliamentary slope ~0.18  |  Presidential slope ~0.08  |  Exact by arithmetic"
  )
```

::: {.callout-note}
## Alternative implementation

```r
# Direct posterior summary of the interaction term
as_draws_df(fit_interact) |>
  summarise(
    mean_b3  = mean(`b_women_parl:reg_type`),
    q2.5     = quantile(`b_women_parl:reg_type`, 0.025),
    q97.5    = quantile(`b_women_parl:reg_type`, 0.975),
    prob_pos = mean(`b_women_parl:reg_type` > 0)
  )
```

:::

**Reading the figure**: the two density curves should be separated.
`P(β₃ > 0 | data)` near 1.0 means we are nearly certain the interaction is
positive — parliamentary systems show a steeper women-in-parliament–health-spending
association.
This is a direct posterior probability statement, not a p-value.

## Step 8: LOO-IC Model Comparison {#sec-part4-step8}

We compare `fit_multi` (no interaction) against `fit_interact` (interaction) using LOO-IC.
Because we generated the data with an interaction, `fit_interact` should dominate `fit_multi`.

```{r}
#| label: loo-interact
#| echo: true
fit_multi <- add_criterion(fit_multi, "loo")
fit_interact <- add_criterion(fit_interact, "loo")
loo_compare(fit_multi, fit_interact)
```

::: {.callout-note}
## Alternative implementation

```r
# WAIC as alternative to LOO-IC — should rank models identically
fit_multi    <- add_criterion(fit_multi,    "waic")
fit_interact <- add_criterion(fit_interact, "waic")
loo_compare(fit_multi, fit_interact, criterion = "waic")
```

:::

**Reading the output**:

- The model listed first has the highest elpd — it is the better predictive model.
- `elpd_diff` = 0 for the winner; all other models show a negative difference
  (i.e., they predict worse).
- When `|elpd_diff| > 2 × se_diff`, the evidence strongly favours the winner.

**Remaining limitations**:

1. The linear interaction assumes the slope *difference* is constant across all
   values of women in parliament — check with a LOESS diagnostic if needed.
2. Country-year observations within the same country are correlated (assumption A5) —
   a multilevel structure (Week 12) would handle this.
3. Omitted confounders (institutional quality, colonial history, cultural factors)
   may bias the women-in-parliament coefficient.

---

# Session Info and Execution Time {#sec-session}

```{r}
#| label: session-info
#| eval: true
sessionInfo()
```

```{r}
#| label: exec-time
#| eval: true
end_time <- Sys.time()
cat(
  "Total execution time:",
  round(difftime(end_time, start_time, units = "mins"), 1),
  "minutes\n"
)
```
