---
title: "Week 9 Lab: Multivariate Regression and Interactions"
subtitle: "Applications from the lecture — work through the code and check that it runs"
author: "Firstname Lastname"
date: last-modified
date-format: MMMM D, YYYY
format: html
toc: true
toc-depth: 3
code-fold: true
code-tools: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# Setup {#sec-setup}

Before running any analysis, we load the packages we need and define a few
shared colour constants that keep plots visually consistent.
`brms` fits Bayesian regression models via Stan; `bayesplot` produces
MCMC diagnostics; `patchwork` assembles multi-panel figures.

```{r}
#| label: setup
#| code-fold: false
start_time <- Sys.time() # used to measure execution time at the end

library(tidyverse) # data manipulation and ggplot2
library(brms) # Bayesian regression via Stan
library(bayesplot) # MCMC diagnostic plots
library(patchwork) # composing multi-panel figures

# Shared colour palette — reused in every plot so figures look consistent
col_1 <- "steelblue"
col_2 <- "#E07B39"
col_3 <- "#7B3F9E"
col_4 <- "goldenrod3"
chain_cols <- c(col_1, col_2, col_3, col_4) # one colour per MCMC chain
```

---

# Part II: Application I — Multivariate `brms` Model {#sec-part2}

## DAGs in Quarto: MermaidJS {#sec-dag-mermaid}

One quick way to sketch a DAG directly in a Quarto document — no R package required.
Use a fenced `{mermaid}` block with `graph LR` (left-to-right) or `graph TD` (top-down).

```{mermaid}
graph LR
  GDP["GDP per Capita"] --> WP["Women in Parliament (%)"]
  GDP --> HS["Health Spending (% GDP)"]
  WP --> HS
```

Nodes take labels in `["quoted text"]`; directed edges are `-->`.
This is enough for a homework submission or a quick theoretical sketch.

## DGP: Synthetic Data (Dataset 1) {#sec-part2-dgp}

Both applications in this lab use **synthetic (simulated) datasets**.
Working with synthetic data is pedagogically powerful: we know the true
parameter values, so we can verify whether our models recover them.
This is something you can almost never do with real data — in practice the DGP
is unobserved; we only see the outcomes, never the process that produced them.

The research question is: *Does the share of women in parliament associate with higher public health expenditure?*

The key complication is **confounding**: wealthier countries simultaneously
have more women in parliament *and* spend more on health.
Failing to account for GDP per capita will over-estimate the
legislative representation association.

```{r}
#| label: dgp1
set.seed(2026) # reproducibility: same data every time
n1 <- 120 # 120 synthetic country-years

# ── Population-level parameters (the ground truth) ──────────────────────────
beta_0 <- 2 # intercept
beta_1_women <- 0.12 # partial association: women in parliament -> health spending
beta_2_gdp <- 0.20 # partial association: GDP per capita -> health spending
sigma_1 <- 1.5 # residual standard deviation

# ── Data-generating process ──────────────────────────────────────────────────
gdp_pc <- rnorm(n1, mean = 18, sd = 7)
women_parl <- 15 + 0.9 * gdp_pc + rnorm(n1, 0, 5)
health_spend <- beta_0 +
  beta_1_women * women_parl +
  beta_2_gdp * gdp_pc +
  rnorm(n1, 0, sigma_1)

df_controls <- tibble(gdp_pc, women_parl, health_spend)
```

::: {.callout-note}
## Alternative implementation

```r
# OLS check: verify OVB analytically without brms
m_biv  <- lm(health_spend ~ women_parl, data = df_controls)
m_full <- lm(health_spend ~ women_parl + gdp_pc, data = df_controls)
round(coef(m_biv)["women_parl"] - coef(m_full)["women_parl"], 3)
```

:::

The true coefficient we want to recover is **$\beta_{\text{women}} = 0.12$**.
Because GDP drives both variables, a bivariate model (no control)
will produce an estimate well above 0.12 — that excess is omitted variable
bias (OVB).

## Step 1: Estimand {#sec-part2-step1}

Before writing any code, we state precisely what we are trying to estimate.
Drawing a DAG before analysis keeps this honest: the estimand follows from
the causal graph, not from what the model happens to produce.

Our DAG (presented in the lecture) shows GDP per capita as a **confounder** —
it has causal arrows into both women's parliamentary representation and health
spending.
A confounder opens a "back-door path" that conflates the association we care
about with the spurious GDP-driven correlation.
Conditioning on GDP per capita — including it as a predictor — closes that path.

**Estimand**: the partial association between women's parliamentary
representation and public health expenditure, adjusting for GDP per capita.

*Adjusting for GDP per capita shifts the estimate toward the true DGP value. Whether the adjustment set is sufficient is a substantive question the data alone cannot answer.*

## Step 2: Data Summary {#sec-part2-step2}

The scatter below shows the raw bivariate association between women's
parliamentary share and health spending.
There is a positive trend, but it conflates the legislative association with
the confounding from GDP.
We will visualise the OVB explicitly in Step 7, after fitting both models.

```{r}
#| label: fig-controls-raw
#| echo: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Raw bivariate scatter of women's parliamentary share against health expenditure. The positive trend conflates the legislative association with the GDP confounder."
df_controls |>
  ggplot(aes(x = women_parl, y = health_spend)) +
  geom_point(colour = col_1, alpha = 0.55, size = 2) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Expenditure (% GDP)",
    title = "Raw Bivariate Association",
    subtitle = "Positive trend — but GDP confounds the relationship"
  )
```

::: {.callout-note}
## Alternative implementation

```r
df_controls |>
  ggplot(aes(x = women_parl, y = health_spend, colour = gdp_pc)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_colour_gradient(low = col_1, high = col_2, name = "GDP pc") +
  theme_bw() +
  labs(title = "Confounding visible: GDP colours the point cloud",
       x = "Women in Parliament (%)", y = "Health Expenditure (% GDP)")
```

:::

## Step 3: Formal Model Specification {#sec-part2-step3}

We write out the full statistical model — likelihood and priors — before
touching `brms`.
This makes our assumptions explicit and contestable.

**Likelihood** (unchanged from Week 8 — still Gaussian):

$$
Y_i \sim \text{Normal}(\mu_i,\, \sigma)
$$

$$
\mu_i = \alpha + \beta_X X_i + \beta_Z Z_i
$$

where $X$ = women in parliament (%), $Z$ = GDP per capita (thousands USD, PPP).

**Priors** (one per unknown parameter — all four must be justified):

$$
\alpha \sim \text{Normal}(10, 5) \qquad \beta_X \sim \text{Normal}(0, 2)
$$

$$
\beta_Z \sim \text{Normal}(0, 2) \qquad \sigma \sim \text{Exponential}(1)
$$

The prior on $\beta_X$ says: *before seeing data, a 1 pp increase in women's
representation is associated with a change of somewhere between −6 and +6 pp
in health spending (the Normal(0, 2) range at ±3σ is [−6, +6]).*
This is a weakly informative prior — broad enough to let the data speak,
narrow enough to down-weight implausible extreme values.

## Step 4: Prior Predictive Check {#sec-part2-step4}

A prior predictive check asks: *if we sample parameter values from the priors
and generate data, do the simulated outcomes look like real-world health
spending?*
Public health expenditure as a share of GDP is typically between 0% and 30%.
Prior-predicted values far outside this range would signal that our priors
are miscalibrated.

```{r}
#| label: fig-prior-ppc-controls
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Prior predictive distribution: 2,000 parameter draws from the joint prior, each used to generate predicted health spending for all 120 observations. The shaded band marks the plausible 0–30% GDP range. Most prior mass falls inside — these priors are defensible."
set.seed(2026)
n_pp <- 2000

# For each of the 2000 draws, compute the linear predictor at every observation
# and add Gaussian noise with the drawn σ.
# rowwise() + mutate(y_sim = list(...)) creates one simulated dataset per draw.
prior_pred_controls <- tibble(
  b0 = rnorm(n_pp, 10, 5), # intercept draws — Normal(10, 5)
  b_wp = rnorm(n_pp, 0, 2), # women_parl slope draws — Normal(0, 2)
  b_gdp = rnorm(n_pp, 0, 2), # gdp_pc slope draws — Normal(0, 2)
  sigma = rexp(n_pp, 1) # residual SD draws — Exponential(1)
) |>
  rowwise() |>
  mutate(
    y_sim = list(
      b0 +
        b_wp * df_controls$women_parl +
        b_gdp * df_controls$gdp_pc +
        rnorm(nrow(df_controls), 0, sigma)
    )
  ) |>
  unnest(y_sim) |>
  ungroup()

prior_pred_controls |>
  ggplot(aes(x = y_sim)) +
  annotate(
    "rect",
    xmin = 0,
    xmax = 30,
    ymin = 0,
    ymax = Inf,
    fill = col_4,
    alpha = 0.12
  ) +
  geom_density(fill = col_1, alpha = 0.55, colour = "white", linewidth = 0.4) +
  geom_vline(
    xintercept = c(0, 30),
    colour = col_4,
    linetype = "dashed",
    linewidth = 0.9
  ) +
  annotate(
    "text",
    x = 15,
    y = Inf,
    vjust = 1.4,
    size = 3.2,
    colour = col_4,
    label = "Plausible empirical range: 0–30% GDP"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Predicted Health Expenditure (% GDP)",
    y = "Prior Predictive Density",
    title = "Prior Predictive Check: Health Spending Model",
    subtitle = "Before seeing data — does the prior generate plausible outcomes?"
  )
```

::: {.callout-note}
## Alternative implementation

```r
fit_prior_only <- brm(
  health_spend ~ women_parl + gdp_pc,
  data = df_controls,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2),  class = "b"),
    prior(exponential(1), class = "sigma")
  ),
  sample_prior = "only",
  chains = 2, iter = 1000, seed = 2026
)
pp_check(fit_prior_only, ndraws = 50)
```

:::

## Step 5: Fit the Model with `brms` {#sec-part2-step5}

The `brm()` call mirrors the formal model written in Step 3.
Note that we specify a separate `prior()` for each coefficient using `coef =`.
This is more precise than applying one flat prior to all slopes, and it
documents our assumptions in the code itself.

`brms` translates the model to Stan (C++), compiles it, and samples 4 parallel
Markov chains.
With `iter = 2000` and `warmup = 1000`, each chain contributes 1,000 post-warmup
draws, giving 4,000 total posterior samples of $(\alpha, \beta_X, \beta_Z, \sigma)$.

We fit **both** the bivariate (no control) and multivariate models here so we
can compare their estimates and assess OVB in Step 7.

```{r}
#| label: fit-controls-biv
#| output: false
# Bivariate model: health spending ~ women in parliament only
# This model omits GDP and will over-estimate the association (OVB)
fit_controls_biv <- brm(
  health_spend ~ women_parl,
  data = df_controls,
  prior = c(
    prior(normal(10, 5), class = "Intercept"), # baseline ~10% GDP
    prior(normal(0, 2), class = "b"), # weakly informative prior
    prior(exponential(1), class = "sigma") # residual SD < 5 almost certainly
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

::: {.callout-note}
## Alternative implementation

```r
# Inspect default prior structure before specifying your own
get_prior(
  health_spend ~ women_parl,
  data   = df_controls,
  family = gaussian()
)
```

:::

```{r}
#| label: fit-controls-multi
#| output: false
# Multivariate model: health spending ~ women in parliament + GDP per capita
# Including GDP closes the back-door path and removes the OVB
fit_controls <- brm(
  health_spend ~ women_parl + gdp_pc,
  data = df_controls,
  prior = c(
    prior(normal(10, 5), class = "Intercept"),
    prior(normal(0, 2), class = "b", coef = "women_parl"), # weakly informative
    prior(normal(0, 2), class = "b", coef = "gdp_pc"), # weakly informative
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2026
)
```

::: {.callout-note}
## Alternative implementation

```r
# Confirm the priors actually applied to the fitted model
prior_summary(fit_controls)
```

:::

## Step 6a: MCMC Diagnostics — Traceplots {#sec-part2-step6a}

Before interpreting any results, we must confirm that MCMC has converged.
Three things to check:

1. **Rhat < 1.01** for every parameter: the chains have mixed — all four
   explored the same region of parameter space.
2. **Bulk_ESS and Tail_ESS > 1,000** for every parameter: we have enough
   effective samples to trust posterior summaries and credible intervals.
3. **Traceplots** look like "hairy caterpillars": each chain wanders randomly
   around the same range with no trends, drifts, or stuck periods.

If any diagnostic fails, stop and investigate before interpreting results.

```{r}
#| label: fig-trace-controls
#| echo: true
#| code-fold: true
#| fig-height: 3.6
#| dpi: 500
#| fig-cap: "Traceplots for the four model parameters. Each colour is one chain. The 'hairy caterpillar' pattern — chains overlapping and mixing rapidly — confirms convergence."
mcmc_trace(
  fit_controls,
  pars = c("b_Intercept", "b_women_parl", "b_gdp_pc", "sigma"),
  facet_args = list(ncol = 2)
) +
  scale_colour_manual(values = chain_cols) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90")
  ) +
  labs(
    title = "MCMC Traceplots: Health Spending Model",
    subtitle = "All chains must mix well before interpreting results"
  )
```

::: {.callout-note}
## Alternative implementation

```r
mcmc_rank_hist(
  fit_controls,
  pars = c("b_Intercept", "b_women_parl", "b_gdp_pc", "sigma")
) +
  labs(title = "Rank histograms: uniform bars confirm good mixing")
```

:::

## Step 6b: Posterior Predictive Check {#sec-part2-step6b}

A posterior predictive check (PPC) asks: *do the model's predictions look like the observed data?*
`pp_check()` draws 50 samples from the posterior predictive distribution
(light blue lines) and overlays them on the observed outcome distribution
(dark blue line).
A well-specified model should produce simulated outcome distributions that
closely envelope the observed one.

```{r}
#| label: fig-ppc-controls
#| echo: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Posterior predictive check for the health spending model. The dark line is the observed distribution; light blue lines are 50 draws from the posterior predictive. The Gaussian model fits well."
pp_check(fit_controls, ndraws = 50) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Posterior Predictive Check: Health Spending Model",
    subtitle = "Light blue: model predictions  |  Dark: observed data"
  )
```

::: {.callout-note}
## Alternative implementation

```r
pp_check(fit_controls, type = "stat_2d", stat = c("mean", "sd")) +
  labs(title = "Joint PPC: posterior mean and SD vs observed")
```

:::

## Step 7: Interpret Results {#sec-part2-step7}

`summary()` returns posterior means (`Estimate`), posterior standard deviations
(`Est.Error`), and 95% credible intervals (`l-95% CI` / `u-95% CI`) for each
parameter.
The `Rhat` column confirms convergence (should be < 1.01 everywhere) and
`Bulk_ESS` / `Tail_ESS` confirm adequate effective sample sizes.

```{r}
#| label: controls-summary
#| echo: true
summary(fit_controls)
```

::: {.callout-note}
## Alternative implementation

```r
as_draws_df(fit_controls) |>
  select(b_women_parl, b_gdp_pc, sigma) |>
  summarise(across(
    everything(),
    list(mean  = mean,
         sd    = sd,
         q2.5  = \(x) quantile(x, 0.025),
         q97.5 = \(x) quantile(x, 0.975))
  ))
```

:::

**What to look for**:

- `b_women_parl` posterior mean should be near 0.12 (the true DGP value).
  Compare this to the bivariate estimate — the difference is the OVB.
- `b_gdp_pc` posterior mean should be near 0.20.
- `sigma` should be near 1.5.

### OVB: Comparing Bivariate and Multivariate Estimates

We now compare the `women_parl` coefficient from both models.
The bivariate estimate (no GDP control) is inflated by OVB;
the multivariate estimate recovers a value close to the true DGP value of 0.12.

```{r}
#| label: fig-ovb-comparison
#| echo: true
#| code-fold: true
#| fig-height: 3.0
#| dpi: 500
#| fig-cap: "OVB visualised: posterior mean and 95% credible interval for the women_parl coefficient from the bivariate (orange) and multivariate (purple) models. The dashed line marks the true DGP value β = 0.12. The gap between the two estimates is the omitted variable bias."
post_biv <- posterior_summary(fit_controls_biv)["b_women_parl", ]
post_multi <- posterior_summary(fit_controls)["b_women_parl", ]

coef_df <- tibble(
  Model = factor(
    c("Bivariate\n(no control)", "Multivariate\n(+ GDP per capita)"),
    levels = c("Bivariate\n(no control)", "Multivariate\n(+ GDP per capita)")
  ),
  Estimate = c(post_biv["Estimate"], post_multi["Estimate"]),
  lo = c(post_biv["Q2.5"], post_multi["Q2.5"]),
  hi = c(post_biv["Q97.5"], post_multi["Q97.5"])
)

coef_df |>
  ggplot(aes(x = Model, y = Estimate)) +
  geom_hline(
    yintercept = 0.12,
    colour = "gray50",
    linetype = "dashed",
    linewidth = 0.9
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi, colour = Model),
    width = 0.12,
    linewidth = 1.1
  ) +
  geom_point(aes(colour = Model), size = 5) +
  scale_colour_manual(values = c(col_2, col_3), guide = "none") +
  annotate(
    "text",
    x = 1.52,
    y = 0.127,
    size = 3.3,
    colour = "gray40",
    label = expression(paste("True ", beta, " = 0.12")),
    hjust = 0
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = NULL,
    y = expression(hat(beta) ~ "(women in parliament -> health spending)"),
    title = "OVB: Bivariate Inflates the Estimate",
    subtitle = "Controlling for GDP brings posterior mean near the true DGP value"
  ) +
  scale_y_continuous(
    limits = c(0, 0.35)
  )
```

::: {.callout-note}
## Alternative implementation

```r
bind_rows(
  as_draws_df(fit_controls_biv) |>
    transmute(b_women_parl, model = "Bivariate"),
  as_draws_df(fit_controls) |>
    transmute(b_women_parl, model = "Multivariate (+GDP)")
) |>
  ggplot(aes(x = b_women_parl, fill = model)) +
  geom_density(alpha = 0.55) +
  geom_vline(xintercept = 0.12, linetype = "dashed", colour = "gray40") +
  scale_fill_manual(values = c(col_2, col_3)) +
  theme_bw() +
  labs(title = "Full posterior densities: OVB as distributional shift",
       x = expression(beta["women"]), fill = NULL)
```

:::

### Visualising the Partial Effect

Rather than just reporting the coefficient, we visualise the estimated
relationship.
We create a prediction grid that varies `women_parl` across its observed
range while holding `gdp_pc` fixed at its sample mean — this isolates the
partial effect of legislative representation.

`posterior_epred()` returns a matrix of posterior draws of $E[Y \mid X, Z]$
for each grid row.
Taking quantiles across draws gives us the posterior mean and 95% credible
interval for the partial effect at each value of `women_parl`.

```{r}
#| label: fig-controls-epred
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Partial effect of women's parliamentary share on health spending, holding GDP per capita at its sample mean. The ribbon is the 95% posterior credible interval."
# Grid: vary women_parl, hold gdp_pc fixed at mean
grid_controls <- tibble(
  women_parl = seq(
    min(df_controls$women_parl),
    max(df_controls$women_parl),
    length.out = 80
  ),
  gdp_pc = mean(df_controls$gdp_pc)
)

# posterior_epred returns a draws × grid matrix of E[Y | X, Z] values
epred_mat <- posterior_epred(fit_controls, newdata = grid_controls)

grid_controls <- grid_controls |>
  mutate(
    lo = apply(epred_mat, 2, quantile, 0.025), # 2.5th percentile across draws
    mid = apply(epred_mat, 2, median), # median across draws
    hi = apply(epred_mat, 2, quantile, 0.975) # 97.5th percentile across draws
  )

df_controls |>
  ggplot(aes(x = women_parl, y = health_spend)) +
  geom_ribbon(
    data = grid_controls,
    aes(x = women_parl, ymin = lo, ymax = hi, y = mid),
    inherit.aes = FALSE,
    fill = col_1,
    alpha = 0.25
  ) +
  geom_line(
    data = grid_controls,
    aes(x = women_parl, y = mid),
    inherit.aes = FALSE,
    colour = col_2,
    linewidth = 1.2
  ) +
  geom_point(colour = col_1, alpha = 0.45, size = 1.8) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "Women in Parliament (%)",
    y = "Health Expenditure (% GDP)",
    title = "Partial Effect: Women in Parliament -> Health Spending",
    subtitle = paste0(
      "GDP per capita held at mean (",
      round(mean(df_controls$gdp_pc), 1),
      " thousand USD)  |  posterior_epred()"
    )
  )
```

::: {.callout-note}
## Alternative implementation

```r
# brms built-in: partial effect marginalised over other predictors
plot(conditional_effects(fit_controls, effects = "women_parl"))
```

:::

## Step 8: LOO-IC Model Comparison {#sec-part2-step8}

Bayesian leave-one-out cross-validation (LOO-IC) estimates how well each model
would predict *new* data — it penalises complexity implicitly.
We compare the bivariate model (no GDP control) against the multivariate model.

`add_criterion()` appends the LOO estimate to each fitted model.
`loo_compare()` ranks models by expected log predictive density (elpd):
higher elpd = better predictive accuracy.
When `|elpd_diff| > 2 × se_diff`, the evidence for the better model is
substantial.

```{r}
#| label: loo-controls-compare
#| echo: true
fit_controls <- add_criterion(fit_controls, "loo")
fit_controls_biv <- add_criterion(fit_controls_biv, "loo")
loo_compare(fit_controls_biv, fit_controls)
```

::: {.callout-note}
## Alternative implementation

```r
# WAIC as alternative information criterion (same ranking expected)
fit_controls     <- add_criterion(fit_controls,     "waic")
fit_controls_biv <- add_criterion(fit_controls_biv, "waic")
loo_compare(fit_controls_biv, fit_controls, criterion = "waic")
```

:::

The multivariate model should dominate: GDP genuinely predicts health spending,
so including it improves out-of-sample accuracy.

**Remaining limitations** to keep in mind:

1. GDP may not be the only confounder — country institutional quality, colonial
   history, and cultural factors might also drive both representation and spending.
2. Country-year observations within the same country are likely correlated —
   a violation of the independence assumption (A5) that Week 12 addresses with
   multilevel models.
3. Linearity is assumed: we should inspect residuals for non-linear patterns.

---

# Part IV: Application II — Interaction `brms` Model {#sec-part4}

## DGP: Synthetic Data (Dataset 2) {#sec-part4-dgp}

The research question is: *Does the association between unemployment and transfer
spending depend on government ideology?*

The **partisan theory of the welfare state** predicts that left-wing governments
respond more aggressively to rising unemployment by raising social transfers.
Right-wing governments are more fiscally restrained.
This is a claim about **association heterogeneity** — the slope of the
unemployment–spending relationship is steeper under left governments.

In real research the DGP is unobserved — the interaction coefficient $\beta_3$
is what we are trying to *discover*, not something we already know.
Here we know it because we generated the data, which lets us check whether the
model recovers the true values.

```{r}
#| label: dgp2
set.seed(42)
n2 <- 150 # 150 synthetic government-years

# ── Population-level parameters (the ground truth) ──────────────────────────
beta_0 <- 10 # intercept (right-government baseline)
beta_1_unemp <- 0.3 # unemployment slope under right government
beta_2_gov <- 2 # intercept shift for left government
beta_3_interact <- 0.9 # extra unemployment slope under left government (the interaction)
sigma_2 <- 2.5 # residual standard deviation

# ── Data-generating process ──────────────────────────────────────────────────
unemployment <- rnorm(n2, mean = 7.5, sd = 2.5)
left_gov <- rbinom(n2, 1, prob = 0.5) # coin flip: equal left/right share

transfer_spend <- beta_0 +
  beta_1_unemp * unemployment +
  beta_2_gov * left_gov +
  beta_3_interact * unemployment * left_gov + # slope shifts up for left govts
  rnorm(n2, 0, sigma_2)

df_interact <- tibble(unemployment, left_gov, transfer_spend) |>
  mutate(government = if_else(left_gov == 1, "Left", "Right"))
```

::: {.callout-note}
## Alternative implementation

```r
# OLS preview: inspect interaction coefficient before running brms
m_ols <- lm(transfer_spend ~ unemployment * left_gov, data = df_interact)
round(coef(m_ols), 3)
```

:::

Key quantities to recover:

| Government | Intercept | Unemployment slope |
|---|---|---|
| Right (`left_gov = 0`) | 10 | **0.3** |
| Left  (`left_gov = 1`) | 12 | **0.3 + 0.9 = 1.2** |

The interaction coefficient **$\beta_3 = 0.9$** captures *how much steeper* the
left slope is relative to the right slope.

## Step 1: DAG and Estimand {#sec-part4-step1}

The partisan welfare state story involves two independent causes of transfer
spending — unemployment and government ideology — where ideology also
*moderates* the unemployment effect.
The DAG has arrows from both Unemployment and Government Ideology into
Transfer Spending; the research question concerns the interaction: is the
unemployment slope steeper under left governments?

Because this DAG has no confounders (no variable that causes both unemployment
and ideology), we do not need a control variable to close a back-door path.
The interaction term itself is the quantity of interest.

**Estimand 1**: $\partial\mu/\partial X|_{Z=0}$ — unemployment slope under
*right* governments.

**Estimand 2**: $\partial\mu/\partial X|_{Z=1}$ — unemployment slope under
*left* governments.

**Estimand 3**: β₃ — the slope difference; how much more does unemployment
associate with spending under left vs. right governments?

## Step 2: Data Summary {#sec-part4-step2}

We start with a quick look at the outcome distribution by ideology group,
then examine the joint scatter and a box-plot comparison.

```{r}
#| label: fig-outcome-dist-interact
#| echo: true
#| fig-height: 3.0
#| dpi: 500
#| fig-cap: "Marginal distribution of transfer spending by government ideology. Left governments show higher average spending — the intercept shift captured by β₂ in the interaction model."
df_interact |>
  ggplot(aes(x = transfer_spend, fill = government, colour = government)) +
  geom_density(alpha = 0.45, linewidth = 0.8) +
  scale_fill_manual(values = c("Left" = col_2, "Right" = col_1)) +
  scale_colour_manual(values = c("Left" = col_2, "Right" = col_1)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Transfer Spending (% GDP)",
    y = "Density",
    title = "Distribution of Transfer Spending by Government Ideology",
    subtitle = "Left governments show higher average spending",
    fill = NULL,
    colour = NULL
  )
```

::: {.callout-note}
## Alternative implementation

```r
df_interact |>
  ggplot(aes(x = government, y = transfer_spend, fill = government)) +
  geom_violin(alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.35, size = 1.2) +
  scale_fill_manual(
    values = c("Left" = col_2, "Right" = col_1), guide = "none"
  ) +
  theme_bw() +
  labs(title = "Transfer spending by ideology — violin view",
       x = NULL, y = "Transfer Spending (% GDP)")
```

:::

The scatter below puts unemployment on the x-axis and draws separate OLS lines
for each ideology group.
If the lines have visibly different slopes, that is preliminary evidence for
an interaction.
The box-plot facets unemployment into terciles and shows the widening gap
between left and right governments as unemployment rises — the visual signature
of β₃ > 0.

```{r}
#| label: fig-interact-summary
#| echo: true
#| code-fold: true
#| fig-height: 3.6
#| dpi: 500
#| fig-cap: "Left: scatter with separate OLS lines per ideology group — slopes look clearly different. Right: transfer spending by unemployment tercile and ideology — the gap widens with unemployment, the visual signature of a positive interaction."
p_scatter_int <- df_interact |>
  ggplot(aes(x = unemployment, y = transfer_spend, colour = government)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(
    aes(group = government),
    method = "lm",
    se = FALSE,
    linewidth = 1.2
  ) +
  scale_colour_manual(
    values = c("Left" = col_2, "Right" = col_1),
    name = "Government"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Unemployment (%)",
    y = "Transfer Spending (% GDP)",
    title = "Separate Slopes by Ideology",
    subtitle = "OLS line per group"
  )

df_interact_plot <- df_interact |>
  mutate(
    unemp_tercile = cut(
      unemployment,
      breaks = quantile(unemployment, c(0, 1 / 3, 2 / 3, 1)),
      labels = c("Low\n(~4%)", "Mid\n(~8%)", "High\n(~12%)"),
      include.lowest = TRUE
    )
  )

p_box_int <- df_interact_plot |>
  ggplot(aes(x = unemp_tercile, y = transfer_spend, fill = government)) +
  geom_boxplot(
    alpha = 0.7,
    position = position_dodge(width = 0.75),
    width = 0.6
  ) +
  scale_fill_manual(
    values = c("Left" = col_2, "Right" = col_1),
    name = "Government"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Unemployment Tercile",
    y = "Transfer Spending (% GDP)",
    title = "Gap Widens with Unemployment",
    subtitle = "Signature of a positive interaction term"
  )

p_scatter_int | p_box_int
```

::: {.callout-note}
## Alternative implementation

```r
df_interact |>
  mutate(tercile = cut(unemployment,
    breaks = quantile(unemployment, c(0, 1/3, 2/3, 1)),
    labels = c("Low unemp.", "Mid unemp.", "High unemp."),
    include.lowest = TRUE)) |>
  ggplot(aes(x = tercile, y = transfer_spend, fill = government)) +
  geom_boxplot(position = position_dodge(0.75), alpha = 0.7) +
  scale_fill_manual(values = c("Left" = col_2, "Right" = col_1)) +
  theme_bw() +
  labs(title = "Spending gap widens with unemployment",
       x = "Unemployment tercile", y = "Transfer Spending (% GDP)", fill = NULL)
```

:::

## Step 3: Formal Model Specification {#sec-part4-step3}

The interaction model augments the additive model with a product term:

$$
Y_i \sim \text{Normal}(\mu_i,\, \sigma)
$$

$$
\mu_i = \alpha + \beta_1 U_i + \beta_2 G_i + \beta_3 (U_i \cdot G_i)
$$

where $U$ = unemployment (%), $G$ = left government (0/1).

The two conditional marginal effects are now *functions* of the moderator:

$$
\frac{\partial\mu}{\partial U} = \beta_1 + \beta_3 G \qquad
\frac{\partial\mu}{\partial G} = \beta_2 + \beta_3 U
$$

**Priors**:

$$
\alpha \sim \text{Normal}(12, 5) \qquad
\beta_1, \beta_2 \sim \text{Normal}(0, 2) \qquad
\beta_3 \sim \text{Normal}(0, 2) \qquad
\sigma \sim \text{Exponential}(0.5)
$$

The prior for $\beta_3$ is centred at zero — our prior expectation is *no moderation*.
The data will update this.
This is appropriately humble: we should not encode strong prior beliefs about
the direction of an interaction before seeing the evidence.

We fit two competing models:

- **M1** (no interaction): assumes the unemployment slope is the same for all
  governments — a testable restriction.
- **M2** (interaction): relaxes this.
  If M2 fits substantially better via LOO-IC, we have evidence that the
  restriction is false.

## Step 4: Prior Predictive Check {#sec-part4-step4}

For the interaction model, we draw regression lines from the joint prior —
one bundle per ideology group.
The check asks: *does the prior allow a wide enough range of slopes, including
no moderation (similar slopes), strong moderation (very different slopes), and
implausible configurations (e.g., enormous negative slopes)?*

```{r}
#| label: fig-prior-ppc-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.4
#| dpi: 500
#| fig-cap: "Prior predictive spaghetti: 150 regression lines drawn from the joint prior, separately for left (orange) and right (blue) governments. The prior is wide but not implausible — it allows no interaction, moderate interaction, and strong interaction."
set.seed(42)
n_pp2 <- 150
unemp_seq <- seq(1.5, 14, length.out = 60)

prior_lines_int <- tibble(
  sim = 1:n_pp2,
  b0 = rnorm(n_pp2, 12, 5), # intercept draws — Normal(12, 5)
  b_u = rnorm(n_pp2, 0, 2), # unemployment slope draws — Normal(0, 2)
  b_g = rnorm(n_pp2, 0, 2), # ideology main effect draws — Normal(0, 2)
  b_int = rnorm(n_pp2, 0, 2) # interaction term draws — Normal(0, 2)
) |>
  expand_grid(unemployment = unemp_seq, left_gov = c(0L, 1L)) |>
  mutate(
    y_hat = b0 +
      b_u * unemployment +
      b_g * left_gov +
      b_int * unemployment * left_gov, # slope shifts by b_int for left govts
    government = if_else(left_gov == 1, "Left", "Right")
  )

prior_lines_int |>
  ggplot(aes(
    x = unemployment,
    y = y_hat,
    group = interaction(sim, government),
    colour = government
  )) +
  geom_line(alpha = 0.12, linewidth = 0.5) +
  scale_colour_manual(
    values = c("Left" = col_2, "Right" = col_1),
    name = "Government"
  ) +
  coord_cartesian(ylim = c(-10, 45)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Unemployment (%)",
    y = "Predicted Transfer Spending (% GDP)",
    title = "Prior Predictive Spaghetti: Interaction Model",
    subtitle = "150 lines per group — prior is non-dogmatic about moderation"
  )
```

## Step 5: Fit Both Models {#sec-part4-step5}

We use the `*` operator in the `brm()` formula.
`unemployment * left_gov` is shorthand for `unemployment + left_gov + unemployment:left_gov`.
Always use `*`, not just the colon form (`:`).
The colon-only form omits the main effects, which forces β₁ = 0 and β₂ = 0 —
usually a false and arbitrary constraint.

```{r}
#| label: fit-interact-m1
#| echo: true
#| output: false
# M1: no interaction — assumes unemployment slope is identical for both ideology groups
fit_m1 <- brm(
  transfer_spend ~ unemployment + left_gov, # additive model: same slope for all
  data = df_interact,
  prior = c(
    prior(normal(12, 5), class = "Intercept"), # baseline spending ~12% GDP
    prior(normal(0, 2), class = "b"), # weakly informative prior on slopes
    prior(exponential(0.5), class = "sigma") # residual SD likely < 5
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 42
)
```

::: {.callout-note}
## Alternative implementation

```r
# Inspect default prior structure for M1
get_prior(
  transfer_spend ~ unemployment + left_gov,
  data   = df_interact,
  family = gaussian()
)
```

:::

```{r}
#| label: fit-interact-m2
#| echo: true
#| output: false
# M2: interaction model — allows the unemployment slope to differ by ideology
# * expands to: unemployment + left_gov + unemployment:left_gov
fit_m2 <- brm(
  transfer_spend ~ unemployment * left_gov,
  data = df_interact,
  prior = c(
    prior(normal(12, 5), class = "Intercept"), # baseline spending ~12% GDP
    prior(normal(0, 2), class = "b"), # weakly informative; centred at zero (no moderation by default)
    prior(exponential(0.5), class = "sigma") # residual SD likely < 5
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 42
)
```

::: {.callout-note}
## Alternative implementation

```r
# Confirm priors actually applied to the interaction model
prior_summary(fit_m2)
```

:::

## Step 6: Diagnostics and Posterior Predictive Check {#sec-part4-step6}

We run the same convergence checks as before.
The traceplots and PPC are shown side-by-side for the interaction model (M2).

```{r}
#| label: fig-diag-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Left: traceplots for the four parameters of M2 — all chains mix well. Right: posterior predictive check — model predictions envelope the observed distribution."
p_trace_m2 <- mcmc_trace(
  fit_m2,
  pars = c(
    "b_Intercept",
    "b_unemployment",
    "b_left_gov",
    "b_unemployment:left_gov"
  ),
  facet_args = list(ncol = 2)
) +
  scale_colour_manual(values = chain_cols) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90"),
    legend.position = "none"
  ) +
  labs(title = "Traceplots: Interaction Model (M2)")

p_ppc_m2 <- pp_check(fit_m2, ndraws = 50) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(title = "Posterior Predictive Check (M2)")

p_trace_m2 | p_ppc_m2
```

::: {.callout-note}
## Alternative implementation

```r
mcmc_rank_hist(
  fit_m2,
  pars = c("b_Intercept", "b_unemployment",
           "b_left_gov", "b_unemployment:left_gov")
) +
  labs(title = "Rank histograms: M2 convergence check")
```

:::

## Step 7: Interpreting Results and Posterior Marginal Effects {#sec-part4-step7}

Inspect the summary output.
The key quantity is `b_unemployment:left_gov` — the interaction coefficient β₃.
Its posterior mean should be near 0.9 (the true DGP value).

```{r}
#| label: interact-summary
#| echo: true
summary(fit_m2)
```

::: {.callout-note}
## Alternative implementation

```r
as_draws_df(fit_m2) |>
  mutate(
    slope_right = b_unemployment,
    slope_left  = b_unemployment + `b_unemployment:left_gov`
  ) |>
  summarise(across(
    c(slope_right, slope_left),
    list(mean  = mean,
         q2.5  = \(x) quantile(x, 0.025),
         q97.5 = \(x) quantile(x, 0.975))
  ))
```

:::

### Posterior Line Clouds

A powerful way to visualise an interaction model is to plot many regression
lines directly from the MCMC draws.
`as_draws_df()` returns a data frame with one row per MCMC draw and one column
per parameter.
For each draw, we compute the intercept and slope for each ideology group
by simple arithmetic and draw the resulting line.

The bundle of blue lines (right governments) should be clearly less steep
than the bundle of orange lines (left governments).

```{r}
#| label: fig-posterior-lines-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.5
#| dpi: 500
#| fig-cap: "Posterior regression line clouds: 200 MCMC draws per ideology group. The orange bundle (left governments) is visibly steeper — posterior evidence for β₃ > 0."
# Sample 200 draws for visual clarity (4000 lines would be too dense)
post_draws_m2 <- as_draws_df(fit_m2) |> slice_sample(n = 200)

df_interact |>
  ggplot(aes(x = unemployment, y = transfer_spend, colour = government)) +
  # Right government lines: intercept = b_Intercept, slope = b_unemployment
  geom_abline(
    data = post_draws_m2,
    aes(intercept = b_Intercept, slope = b_unemployment),
    colour = col_1,
    alpha = 0.05,
    linewidth = 0.4
  ) +
  # Left government lines: intercept = b_Intercept + b_left_gov,
  #                          slope   = b_unemployment + b_unemployment:left_gov
  geom_abline(
    data = post_draws_m2,
    aes(
      intercept = b_Intercept + b_left_gov,
      slope = b_unemployment + `b_unemployment:left_gov`
    ),
    colour = col_2,
    alpha = 0.05,
    linewidth = 0.4
  ) +
  geom_point(alpha = 0.5, size = 1.8) +
  scale_colour_manual(
    values = c("Left" = col_2, "Right" = col_1),
    name = "Government"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = "Unemployment (%)",
    y = "Transfer Spending (% GDP)",
    title = "Posterior Line Clouds by Government Ideology",
    subtitle = "200 MCMC draws per group — orange bundle (left) is steeper"
  )
```

::: {.callout-note}
## Alternative implementation

```r
# brms built-in: automatic interaction plot
plot(conditional_effects(
  fit_m2,
  effects = "unemployment:left_gov"
))
```

:::

### Posterior Conditional Marginal Effects

The conditional marginal effect of unemployment for a given ideology group
is a function of β₁ and β₃:

$$
\text{ME}_{\text{right}} = \beta_1 \qquad
\text{ME}_{\text{left}}  = \beta_1 + \beta_3
$$

Because we have full posterior draws of both β₁ and β₃, we compute these
by arithmetic — no approximation or delta method needed.
The resulting collections of values *are* the posterior distributions of the
conditional marginal effects.

```{r}
#| label: fig-marginal-effects-interact
#| echo: true
#| code-fold: true
#| fig-height: 3.4
#| dpi: 500
#| fig-cap: "Posterior distributions of the conditional marginal effect of unemployment on transfer spending by ideology group. The left slope (~1.2) and right slope (~0.3) are well separated — strong posterior evidence for moderation."
post_all_m2 <- as_draws_df(fit_m2)

me_df <- post_all_m2 |>
  transmute(
    `Right Government` = b_unemployment, # ME at G = 0
    `Left Government` = b_unemployment + `b_unemployment:left_gov` # ME at G = 1
  ) |>
  pivot_longer(
    everything(),
    names_to = "government",
    values_to = "marginal_effect"
  ) |>
  mutate(
    government = factor(
      government,
      levels = c("Right Government", "Left Government")
    )
  )

# Compute P(β₃ > 0 | data) directly from posterior draws — a direct probability statement
prob_b3_pos <- mean(post_all_m2$`b_unemployment:left_gov` > 0)

me_df |>
  ggplot(aes(x = marginal_effect, fill = government, colour = government)) +
  geom_density(alpha = 0.55, linewidth = 0.8) +
  geom_vline(
    xintercept = 0,
    colour = "gray40",
    linetype = "dotted",
    linewidth = 0.9
  ) +
  scale_fill_manual(
    values = c("Right Government" = col_1, "Left Government" = col_2),
    name = NULL
  ) +
  scale_colour_manual(
    values = c("Right Government" = col_1, "Left Government" = col_2),
    name = NULL
  ) +
  annotate(
    "text",
    x = 1.7,
    y = 1.2,
    size = 3.5,
    colour = col_3,
    label = paste0("P(β₃ > 0 | data) = ", round(prob_b3_pos, 3))
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    x = expression(partialdiff * mu / partialdiff ~ Unemployment),
    y = "Posterior Density",
    title = "Posterior Marginal Effects: Unemployment -> Transfer Spending",
    subtitle = "Left slope ~1.2  |  Right slope ~0.3  |  Exact by arithmetic"
  )
```

::: {.callout-note}
## Alternative implementation

```r
# Direct posterior summary of the interaction term
as_draws_df(fit_m2) |>
  summarise(
    mean_b3  = mean(`b_unemployment:left_gov`),
    q2.5     = quantile(`b_unemployment:left_gov`, 0.025),
    q97.5    = quantile(`b_unemployment:left_gov`, 0.975),
    prob_pos = mean(`b_unemployment:left_gov` > 0)
  )
```

:::

**Reading the figure**: the two density curves should be well separated.
`P(β₃ > 0 | data)` near 1.0 means we are nearly certain the interaction is
positive — left governments show a steeper unemployment–spending slope.
This is the Bayesian analogue of a one-sided test, but it is a direct
probability statement, not a p-value.

## Step 8: LOO-IC Model Comparison {#sec-part4-step8}

We compare M1 (no interaction) against M2 (interaction) using LOO-IC.
Because we generated the data with an interaction, M2 should dominate M1.

```{r}
#| label: loo-interact
#| echo: true
fit_m1 <- add_criterion(fit_m1, "loo")
fit_m2 <- add_criterion(fit_m2, "loo")
loo_compare(fit_m1, fit_m2)
```

::: {.callout-note}
## Alternative implementation

```r
# WAIC as alternative to LOO-IC — should rank models identically
fit_m1 <- add_criterion(fit_m1, "waic")
fit_m2 <- add_criterion(fit_m2, "waic")
loo_compare(fit_m1, fit_m2, criterion = "waic")
```

:::

**Reading the output**:

- The model listed first has the highest elpd — it is the better predictive model.
- `elpd_diff` = 0 for the winner; all other models show a negative difference
  (i.e., they predict worse).
- When `|elpd_diff| > 2 × se_diff`, the evidence strongly favours the winner.

**Remaining limitations**:

1. The linear interaction assumes the slope *difference* is constant across all
   unemployment levels — check with a LOESS diagnostic when Z is continuous.
2. Government-years within the same country are correlated (A5) — a multilevel
   structure (Week 12) would handle this.
3. Omitted confounders (economic openness, union density) may bias the ideology
   coefficient.

---

# Session Info and Execution Time {#sec-session}

```{r}
#| label: session-info
#| eval: true
sessionInfo()
```

```{r}
#| label: exec-time
#| eval: true
end_time <- Sys.time()
cat(
  "Total execution time:",
  round(difftime(end_time, start_time, units = "mins"), 1),
  "minutes\n"
)
```
